---
title: "Analysing scales"
author: "Paul Sharpe, Andy Wills, Sophie Homer"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Data required to knit
## https://github.com/ajwills72/rminr-data/tree/master/going-further/sses.csv
##
## I check out rminr-data and make a symbolic link to going-further

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
options(tibble.width = Inf) # show all columns in output
```

# Contents

- [Introduction](#intro)

- [Getting started](#start)

- [Calculating Cronbach’s alpha](#cronbach)

<a name="intro"></a>

# Introduction

This worksheet describes analyses you are likely to need when working with pscyhometric scales.

<a name="start"></a>

# Getting started

To prepare for this worksheet:

1. Open the project you used to complete the [Preprocessing scales worksheet](preproc-scales.html).
1. Add the code below to the end of `scales.R`.

<a name="cronbach"></a>

# Calculating Cronbach’s alpha

Cronbach's alpha measures the internal consistency of a scale, using correlations between different groupings of individual scale items. Values of alpha range from 0-1, with higher numbers indicating a more reliable scale. When a new scale is developed, its reliability is normally reported using Cronbach's alpha. It's wise to ensure that the alpha for your sample data is similar to this value.

We'll demonstrate the calculation of Cronbach's alpha for the State Self-Esteem Scale (SSES), which was introduced in the [Preprocessing scales worksheet](preproc-scales.html). As you will recall, this experiment used the SSES to measure self-esteem before and after a mental imagery intervention. This was a 2 (time) x 3 (condition) mixed design, with the following experimental conditions:

 * Condition 1: participants visualised a negative mental image of themself
 * Condition 2: participants visualised a negative mental image of someone else
 * Condition 0 (control): participants did a card sorting task and did not think of any images.

We'll write a function to calculate alpha for the SSES using the `cronbach()` function from the `psy` package. You may need to install this package before you can use it for the first time -- if so, use `install.packages('psy')`.

```{r cronbach}
library(psy)

sses_cronbach <- function(scores) {
  cronbach <- cronbach(select(scores, q1:q20))
  data.frame(cronbach$alpha)
}
```

**Explanation of commands:**

1. Our function `sses_cronbach` expects its `scores` argument to be a data frame containing SSES scores.

1. The `cronbach()` function expects a wide data frame (one row per subject), with columns containing the scale scores. We supply this using the function `select(scores, q1:q20)`. Notice how we can use the column names in `scores` to specify the range of columns to `select()`.

1. In addition to the value of alpha, `cronbach()` returns the number of subjects and the number of items. Here were only interested in alpha, which we return in a data frame. Returning alpha in a data frame will help us to use the function for calculating alpha for each condition in our data.

We can now calculate an overall alpha for SSES by calling `sses_cronbach` with all of our SSES data, using `bind_rows(sses_pre, sses_post)`:

```{r cronbach-all}
sses_cronbach_total <- sses_cronbach(bind_rows(sses_pre, sses_post))
sses_cronbach_total
```

This gives us a value of `0.67`, which is quite a bit lower than the `0.92` reported by the authors who developed the SSES (Heatherton & Polivy, 1991).

We can use the same function to calculate alpha for the pre-intervention SSES scores for each condition in our experiment:

```{r cronbach-pre}
sses_pre_cronbach <- sses_pre %>%
  group_by(condition) %>%
  group_modify(~ sses_cronbach(.x))
sses_pre_cronbach
```

**Explanation of pipeline commands:**

1. We start by piping `sses_pre` into `group_by(condition)`.

1. We pipe this grouped data frame into `group_modify(~ sses_cronbach(.x))`. The `group_modify()` function allows you to run a function for each group in a grouped data frame. The `~` tells `group_modify()` to treat the next argument as a function, in this case `sses_cronbach(.x)`. The `.x` means the subset of rows created by our `group_by()`. i.e. the SSES scores for each experimental condition. The function used within `group_modify()` must return its result as a data frame, which explains why we returned alpha as a data frame in `sses_cronbach()`. So this code runs `sses_cronbach()` once for each set of rows that makes up our three conditions, with the three values of alpha combined into a single data frame.

1. The output of the pipeline is stored in `sses_pre_cronbach`. 

You can see that `sses_pre_cronbach` has a `condition` column, and a column containing Cronbach's alpha for each condition.

We can use a similar pipeline to calculate post-intervention alphas:

```{r cronbach-post}
sses_post_cronbach <- sses_post %>%
  group_by(condition) %>%
  group_modify(~ sses_cronbach(.x))
sses_post_cronbach
```

The pre and post SSES data show that alpha is especially low in the `self` and `other` conditions.

<a name="baseline"></a>
___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

