---
title: "Better graphs"
author: "Andy Wills"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA)

```

# Contents

- [Introduction](#intro)

- [Meaningful labels](#labels)

- [Journal styling](#style)

- [High-quality output](#hq)

<a name = "intro"></a>

# Introduction

In this worksheet, we'll look at how to produce publication-quality graphs in R, using an example from a previous worksheet.

Specifically, in the [within-subject differences](anova1.html#densediff) worksheet, we produced a density plot of the within-subject differences in reaction time for congruent versus incongruent trials

```{r init, message=FALSE, echo=FALSE}
library(tidyverse)
words <- read_csv("wordnaming2.csv")
wordctrl <- words %>% filter(medit == "control")
wordctrlCI <- wordctrl %>% filter(congru != "neutral")
wordctrlCIsum <- wordctrlCI %>% group_by(subj, congru) %>% summarise(rt = mean(rt))
ctrl <- wordctrlCIsum %>% pivot_wider(names_from = congru, values_from = rt)
ctrldiff <- ctrl %>% mutate(diff = incong - cong)
```

```{r graph}
ctrldiff %>% ggplot(aes(diff)) +
    geom_density(aes(y=..scaled..)) +
    geom_vline(xintercept = 0, colour = 'red')
```

This graph looks OK, but would need some improvement before including in a report or journal article. Here are the steps of this makeover.

<a name = "labels"></a>

# Meaningful labels

The first thing to do is change the axis labels to something a bit more human readable. We use the `xlab` and `ylab` commands for this. We covered these commands previously in the [Absolute Beginners' guide](exploring-incomes.html#custom-graphs):

```{r graph2}
ctrldiff %>% ggplot(aes(diff)) +
    geom_density(aes(y=..scaled..)) +
    geom_vline(xintercept = 0, colour = 'red') +
    xlab("Incongruent RT - Congruent RT (ms)") +
    ylab("Scaled density")
```

<a name = "style"></a>

# Journal styling

The default styling for `ggplot` is different to what is preferred in most psychology journals. Fortunately, we can use Tina Seabrooke's `theme_APA` to correct this. You'll find the code in the same _git_ repository as the data, so all you need to do is load in her code:

```{r source}
source("themeapa.R")
```

and then add it as a theme to your graph (much as you have used `theme_bw` in the past):

```{r graph3}
ctrldiff %>% ggplot(aes(diff)) +
    geom_density(aes(y=..scaled..)) +
    geom_vline(xintercept = 0, colour = 'red') +
    xlab("Incongruent RT - Congruent RT (ms)") +
    ylab("Scaled density") +
    theme_APA
```

<a name = "hq"></a>

# High-quality output

If you are writing a report or journal article, it's generally a bad idea to screenshot your graphs, and it's also generally a bad idea to use the _export_ functionality within Rstudio. This is because both of these options produce graphics that are not high enough quality for publication.

To produce high-quality output, you should first create an object for your graph, much as we created an [object for the output of analysis](anova3.html#bfact).

```{r graph4}
dgraph <- ctrldiff %>% ggplot(aes(diff)) +
    geom_density(aes(y=..scaled..)) +
    geom_vline(xintercept = 0, colour = 'red') +
    xlab("Incongruent RT - Congruent RT (ms)") +
    ylab("Scaled density") +
    theme_APA
dgraph
```

We can now use the `ggsave` command to save a high-quality version of that graph:

```{r ggsave}
ggsave(filename = "fig1.pdf", plot = dgraph, units = "cm", width = 15, height = 10)
```

## Explanation of command

`filename = "fig1.pdf"` - Save the graph as _fig1.pdf_. Try to use PDF where possible, because it produces the best quality output and the smallest file size. However, if you're unfortunate enough to be using a wordprocessor than cannot import PDF graphs (e.g. Microsoft Word) then you can use PNG format instead. You do this by changing the filename, e.g. `filename = "fig1.png"`. If you send your paper to a journal for consideration, they will also require the PDF version of your graphs as a separate attachement, as PNG files are generally not good enough for professional publications. For most internal reports (and university coursework), PNG is generally good enough.

`plot = dgraph` - Use the object called `dgraph` as the graph you want to save.

`units = "cm"` - The following commands will set the size of the graph; this command says what units these are in. You usually want to use "cm" (centimetres) but if you live in a country that hasn't adopted the metric system, you can use "in" (inches) instead. 

`width = 15` - The graph (including border etc.) should be 15 units wide (the units in this case being centimetres)

`height = 10` - The graph (including border etc.) should be 10 units high (centimetres in this case). 

## Explanation of output

A file called _fig1.pdf_ will have appeared in your _Files_ window in RStudio. You can export this in the [usual way](using-projects.html#download).

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 


