---
title: "Better tables"
author: "Paul Sharpe, Andy Wills, Allegra Cattani"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Data required to knit
## https://github.com/ajwills72/rminr-data/tree/master/going-further/picture-naming.csv
## https://github.com/ajwills72/rminr-data/tree/master/case-studies/jon-may/big5_total.csv
##
## I check out rminr-data and make a symbolic link to going-further

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
library(pander)
library(tidyverse)
```

## Contents

- [Introduction](#intro)

- [Getting started](#start)

- [Creating a correlation matrix](#cor-matrix)

- [Exercise 1](#ex1)

- [Creating a custom table of descriptive statistics](#descriptives)

- [APA-format tables](#apa)

- [Copying to your wordprocessor](#kable)

- [Exercise 2](#ex2)

- [R Markdown](#rmd)

<a name="intro"></a>

## Introduction

It can be helpful to present data in tables, rather than text, especially when you need to refer to the same data in different parts of a report. Although tables can be produced manually using a word processor, generating them directly from your data ensures they are up-to-date, and reduces copy-paste errors. This worksheet explains how to use `R` to produce some of the types of table used to report psychological research.

<a name="start"></a>

## Getting started

To prepare for this worksheet:

1. Open the `rminr-data` project we used [previously](preproc.html#load).

1. If you don't see a folder named `going-further`, it means you created your project _before_ the data required for this worksheet was added to the `rminr-data` git repository. You can get the latest files by asking git to "`pull`" the repository. Select the `Git` tab, which is located in the row of tabs which includes the `Environment` tab. Click the `Pull` button with a downward pointing arrow. A window will open showing the files which have been pulled from the repository. Close the `Git pull` window.

1. Open the `Files` tab. The `going-further` folder should contain the file `picture-naming-preproc.csv`.

1. Create a script named `tables.R` in the `rminr-data` folder (the folder above `going-further`). Add the code to this script as you work through each section of the worksheet.

<a name="cor-matrix"></a>

## Creating a correlation matrix

We'll start by producing a correlation matrix. A correlation matrix shows correlations between all combinations of a set of variables, which is often required in research reports. We'll demonstrate an easy way to produce correlation matrices, with [APA styling](https://apastyle.apa.org/style-grammar-guidelines/tables-figures/tables), in a format that can be read by Microsoft Word or LibreOffice Writer. A similar approach can be used to produce other common table types.

We'll generate a correlation matrix using the `attitude` dataset, which is included with `R`. These data are the percentage of favourable attitudes given by employees, in relation to seven questions regarding their department (you can find out a bit more about these data by typing `?attitude`). Here are the first few rows of the data frame:

```{r, echo=FALSE}
attitude %>% head(5) %>% pander()
```

We'll use the `apaTables` package to generate the correlation matrix.

```{r, R.options = list(width = 10000)}
rm(list = ls()) # clear the environment
library(apaTables)
apa.cor.table(attitude, filename="table1.doc", table.number = 1)
```

**Explanation of commands:**

We load the `apaTables` package.The function to generate a correlation matrix is `apa.cor.table()`. We pass the `attitude` data frame as the first argument, and use `filename` to specify that the output should be saved in the file `table1.doc`. The `table.number` argument sets the number in the table heading output, in this case "Table 1". If you omit this argument, the text will be "Table XX".

**Explanation of output:**

[Export](https://ajwills72.github.io/rminr/using-projects.html#download) `table1.doc` from RStudio and open it using a word processor

The first thing to notice is that the styling (spacing, use of italics, horizontal lines, positioning of captions and footnotes etc.) complies with the APA guidelines for tables. 

The table number and caption is above the table itself - you will need to edit the caption by hand to make it more meaningful, for example "Means, standard deviations, and correlations with confidence intervals, for the attitude measures of Study 1".

The `Variable` column contains a number and the column name for the seven attitude variables. The next two columns show the mean and standard deviation for each variable. The remaining columns use the numbers from items in the `Variable` column as headings, indicating that they refer to the same variable. The cells show the correlation between the column variables and each of the variables in the rows. Cells are left empty where a variable would otherwise be correlated with itself. The 95% [confidence interval](https://ajwills72.github.io/rminr/more-on-t.html) for the correlation is shown in square brackets.

For example, the correlation between `rating` and `complaints` in this sample is .83. The confidence interval indicates that the population value is likely to be between .66 and .91. 

Evidence for the correlation is calculated using traditional statistics, rather than the Bayes factors described in the [Relationships, part 2 worksheet](corr.html). One asterisk (`*`) indicates `p < .05`. Two asterisks (`**`) signify `p < .01`. These calculations assumed a two-tailed test; one-tailed tests for correlations are explained in the [More on relationships, part 2](corr-extended.html) worksheet. Also recall that p-values are [widely misinterpreted](evidence.html#p-wrong), so it would be better to edit this part of the table by hand to reflect [Bayes Factors](corr.html#evidence) you have already calculated. We suggest using `*` for BF > 3, `**` for BF > 10, `o` for BF < 0.33, and `oo` for BF < 0.1. Change the text at the bottom of the table accordingly.

<a name="ex1"></a>

## Exercise 1

For this exercise, we'll load some data from a study which measured aspects of participants' personality.

```{r}
# Exercise 1
library(tidyverse)
big5 <- read_csv('case-studies/jon-may/big5_total.csv')
```

The first few rows show that the scale used measured the 'big 5' personality factors; openness to experience, conscientiousness, extraversion, agreeableness and neuroticism (OCEAN).

```{r, echo=FALSE}
big5 %>% head(5) %>% pander()
```

Create a correlation matrix for the five personality factors. Number the table as "Table 2", and save the results in `table2.doc`. Your table should look like this in Rstudio:

```{r ex1, echo=FALSE, R.options = list(width = 10000)}
apa.cor.table(big5 %>% select(-subj), filename='table2.doc', table.number = 2)
```

...and it should be APA formatted in the file `table2.doc`. 

**Copy the R code you used for this exercise into PsycEL**

<a name="descriptives"></a>

## Creating a custom table of descriptive statistics

As with graphs, there is often an element of design involved in presenting tabular data in a format most useful for your reader. Packages like `apaTables` are useful for producing APA tables where there is a standard way to present data. However, you often need a table which is customised to present your data in the most useful format. The cost of custom tables is that the content requires a little more preprocessing, and styling the table according to APA standards will require some hand-formatting in your wordprocessor.

We'll demonstrate this process by producing a table of descriptive statistics. The data we'll use comes from an experiment which evaluated children’s language development using the Words in Game (WinG) test. WinG consists of a set of picture cards which are used in four tests: noun comprehension, noun production, predicate comprehension, and predicate production. The Italian and English versions of the WinG cards use different pictures to depict the associated words. The experiment tested whether English-speaking children aged approximately 30 months, produce similar responses for the two sets of cards. We would like to produce a single table, containing descriptive statistics for all four tests.

We start by loading the data:

```{r excel, message=FALSE}
# Load data
wing_preproc <- read_csv('going-further/picture-naming-preproc.csv')
```

The first few rows of `wing_preproc` look like this:

```{r, echo=FALSE}
wing_preproc %>% head() %>% pander(split.table = Inf)
```

### Table of descriptives

Our test scores are currently in _wide_ format (lots of columns, few rows), but R generally requires data to be in _long_ format 
(lots of rows, few columns). This means we first have to make the data frame wider, so we can calculate summary statistics.

```{r pivot}
# wide to long
task_by_subj <- wing_preproc %>%
  pivot_longer(cols = c(nc, np, pc, pp),
               names_to = 'task',
               values_to = 'correct') %>%
  select(subj, gender, cards, task, correct)
```

**Explanation of command:**

In the [Within-subject differences worksheet](anova1.html#pivot), you learned how to use `pivot_wider()` to widen long data frames. The `pivot_longer()` command does the reverse -- it lengthens wide data frames. `cols = c(nc, np, pc, pp)` selects the columns we want to pivot. Each value in these columns is added to a row in a new column called `correct` (`values_to = 'correct'`). In the same row, a new column `task` is set to the name of the column which the value came from (`names_to = 'task'`). All of the values in the other columns are duplicated for each row. We [select](preproc.html#select) just the columns we want for our table of descriptive statistics.

The first few rows of `task_by_subj` look like this:

```{r echo=FALSE}
task_by_subj %>% head(5) %>% pander()
```

Now we can calculate some summary statistics, using commands that we've already used in previous worksheets:

```{r descriptives-1}
# Table of descriptive statistics

descript <- task_by_subj %>%
  group_by(task, gender) %>%
  summarise(mean = mean(correct, na.rm = TRUE), sd = sd(correct, na.rm = TRUE))
```

**Explanation of commands:**

1. We've come across `group_by` [before](group-differences.html#group), here we use it to group the data by two variables at the same time, `task` and `gender`, giving us eight groups overall. 

2. We've also come across `summarize` [before](exploring-incomes.html#means), including the use of `na.rm = TRUE` to deal with missing data. 


Our data now looks like this:

```{r echo=FALSE}
descript %>% pander()
```

### Meaningful labels

The `descript` data frame contains just the numbers we want to include in our report - the means and standard deviations for each of the eight groups. However, the row labels (`np`, etc.) are not particularly clear, so we replace them with something more human readable:


```{r mean-labels}
task_names <- c(
  nc = 'Noun Comprehension',
  np = 'Noun Production',
  pc = 'Predicate Comprehension',
  pp = 'Predicate Production'  
)

descript$task <- descript$task %>% recode(!!!task_names)
```

**Explanation of commands**: We're using the `recode` command that we've previously used in the [cleaning up questionnaire data](https://benwhalley.github.io/rmip/worksheet-recoding.html) worksheet:

1. We start by telling R what each of the codes, `nc` etc., mean. So, for example `nc = 'Noun Comprehension'`. We combine the four 'translations' together into `task_names` using `c()` (short for 'concatenate', i.e. put things together).

2. We then take the `task` columns of the `descript` data frame (`descript$task`) and pipe (`%>%`) it to `recode`, where it uses `task_names` to do the recoding. We write (`<-`) that result back into `descript$task`. 

3. It's not critical to understand the `!!!` part of the `recode` command, but an explanation is [here](https://benwhalley.github.io/rmip/explain-exclamationmarks.html#explain-exclamationmarks) if you are curious.

Our table now looks like this:

```{r echo=FALSE}
descript %>% pander()
```

<a name="apa"></a>

## APA-format tables

Our table is now clear and easy to read. We could include it in a report without much further effort, and the reader would be able to easily see what we wanted to show them. However, it is not quite in the format that psychologists are most familiar with (which is [APA format](https://apastyle.apa.org/style-grammar-guidelines/tables-figures/sample-tables#tests)). In APA format, the table would look more like this: 

```{r echo=FALSE}
# Format wider
descript_table <- descript %>%
  pivot_wider(names_from = gender, values_from = c(mean, sd))

# Order columns
descript_table <- descript_table %>% select(task, mean_female, sd_female, mean_male, sd_male) 

# Column names
colnames(descript_table) <- c("Task", "Female (M)", "Female (SD)", "Male (M)", "Male (SD)")

# Rounding
descript_table[,2:5] <- round(descript_table[,2:5], 2)

#Output
descript_table %>% pander()

# Clear vars
rm(descript_table)
```

In other words, it would be _wider_: more columns and fewer rows. 

We can widen the table, using the `pivot_wider` command we have previously used in the [within-subject differences](anova1.html#pivot) worksheet:

```{r descriptives-2}
# Widen table
descript_table <- descript %>%
  pivot_wider(names_from = gender, values_from = c(mean, sd)) 
```

Our table now has the same format as an APA table...

```{r echo=FALSE}
descript_table %>% pander()
```

...but the columns are in a different order. APA format dictates that means should be placed next to their associated standard deviations in a table (APA format is weirdly specific). Fortunately, we can rearrange columns using the `select` command that we've come across before:

```{r rearrange}
# Re-order columns
descript_table <- descript_table %>% select(task, mean_female, sd_female, mean_male, sd_male) 
```

```{r echo=FALSE}
descript_table %>% pander()
```

Finally, we can replace the column names with something a bit more human readable, using the `colnames` function:

```{r rename-cols}
# Column names
colnames(descript_table) <- c("Task", "Female (M)", "Female (SD)", "Male (M)", "Male (SD)")
```

```{r echo=FALSE}
descript_table %>% pander()
```

Note that it would arguably be clearer to write "mean" rather than "M", but it's another quirk of APA style that we write "M" to stand for mean. 

<a name="kable"></a>

## Copying into your wordprocessor

There are a number of different ways to get a table in R into your wordprocessor. We're going to use the `kableExtra` package, because it's really flexible, so it's capable of producing almost any table you might need. We're only going to use it in the most basic way here; for some other examples of what it can do, see the kableExtra [website](https://haozhu233.github.io/kableExtra/awesome_table_in_html.html).

To get a version of `descript_table` that you can cut-and-paste into your wordprocessor, do this:

```{r kable-descriptives,results='hide'}
library(kableExtra)
descript_table %>% kable(digits=2) %>%  kable_styling()
```

**Explanation of commands:**

1. `library(kableExtra)` loads the `kableExtra` package.
1. We pipe our data into `kable()`. The `digits=2` part ensures that every number is reported to two decimal places.
1. We then pipe `kable()` into `kable_styling()`. This command prints the table to the Viewer window in RStudio.

**Explanation of output:**

Try copying the table into your word processor now. In the **Viewer** pane, select all of the rows and columns in the table, then right-click and select `Copy`. Open your word processor and select `Paste`.

<a name="ex2"></a>

## Exercise 2

Starting with the data in `task_by_subj`, generate a table of descriptive statistics showing task accuracy for the Italian and English cards. It should look like this:

```{r ex2-1, echo=FALSE, message=FALSE}
# Summarise
descriptives_by_cards <- task_by_subj %>%
  group_by(task, cards) %>%
  summarise(mean = mean(correct, na.rm = TRUE), sd = sd(correct, na.rm = TRUE)) 

# Labels
task_names <- c(
  nc = 'Noun Comprehension',
  np = 'Noun Production',
  pc = 'Predicate Comprehension',
  pp = 'Predicate Production'  
)

descriptives_by_cards$task <- descriptives_by_cards$task %>% recode(!!!task_names)

# Widen
by_cards_table <- descriptives_by_cards %>%
  pivot_wider(names_from = cards, values_from = c(mean, sd))

# Order columns
by_cards_table <- by_cards_table %>% select(task, mean_english, sd_english, mean_italian, sd_italian) 

# Column names
colnames(by_cards_table) <- c("Task", "English (M)", "English (SD)", "Italian (M)", "Italian (SD)")

#Output
by_cards_table %>% kable(digits=2) %>% kable_styling()
```

**Copy the R code you used for this exercise into PsycEL.**

<a name="rmd"></a>

## R Markdown

You can avoid copy-pasting tables (and all other analyses) by writing your reports using `R Markdown` instead of a word processor. `R Markdown` is a language for writing documents which include `R` code. The code is run, and the output is included in the document. `R Markdown` can be used to produce different types of document (e.g. reports, presentations, web pages), in various formats (e.g. Microsoft Word, PDF, HTML). The `Research Methods in R` worksheets are written using `R Markdown`, and although we don't teach it in these materials, there are other [courses](https://rmarkdown.rstudio.com/lesson-1.html) which make it easy to learn.

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

