---
title: "Socioeconomic factors in depression, anxiety and stress"
author: "Alyson Norman, Chelsea Hill, Paul Sharpe"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate
## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
options(tibble.width = Inf) # show all columns in output
```

# Contents

<a name="intro"></a>

## Before you start

This is an advanced worksheet, which assumes you have completed the [Absolute Beginners' Guide to R](https://ajwills72.github.io/rminr/#beginners) course, the [Research Methods in Practice (Quantitative section)](https://benwhalley.github.io/rmip/overview-quantitative.html) course, and the [Intermediate Guide to R](https://ajwills72.github.io/rminr/#rmip) course.

# Introduction

<a name="start"></a>

# Getting started

Open the `rminr-data` project we used [previously](preproc.html#load).

Ensure you have the latest files by asking git to "`pull`" the repository. Select the `Git` tab, which is located in the row of tabs which includes the `Environment` tab. Click the `Pull` button with a downward pointing arrow. A window will open showing the files which have been pulled from the repository. Close the `Git pull` window. The `case-studies` folder should contain a folder named `alison-norman`.

Next, create a new, empty R script and save it in the `rminr-data` folder as `cs-alison-norman.R`. Put all the commands from this worksheet into this file, and run them from there. Save your script regularly.

```{r init-load, message=FALSE}
rm(list = ls()) # clear the environment
library(tidyverse)
```

<a name="preprocessing"></a>

# Preprocessing

See also: [preprocessing scales](preproc-scales.html), [analysing scales](analyse-scales.html).

```{r preprocess-sses-pre}
data <- read_csv('case-studies/alison-norman/ses-dass.csv') %>%
  set_names(~ str_to_lower(.) %>%
              str_replace_all('overall_', '')) %>%
  mutate(ses = factor(ses), ses_category = factor(ses_category),
         ses_level = factor(ses_level))
data$subj <- factor(seq.int(nrow(data)))

anxiety <- data %>%
  select(subj, ses, ses_category, contains('anxiety'))
anxiety <- mutate(anxiety, total = rowSums(anxiety[4:17]))

depression <- data %>%
  select(subj, ses, ses_category, contains('depression'))
depression <- mutate(depression, total = rowSums(depression[4:17]))

stress <- data %>%
  select(subj, ses, ses_category, contains('stress'))
stress <- mutate(stress, total = rowSums(stress[4:17]))
```

TODO: Calculate low, medium, high SES categories

### Notes from Alison on defining SES

* Education either by own or parents. If in education still, must use parents:
  * Q48-b if answer = 0 then use parents education (Q49 + Q50)/2: 0-2 = low, 3-4 = medium, 5-6 = high
  * Q48-b = 1 then use own education (Q48): 0-2 = low, 3-4 = medium, 5-6 = high

* Employment
  * Q48-b if answer = 0 then use parents combined employment (combined SES): 0-2 = low, 3-4 = medium, 5-6 = high
  * Q48-b = 1 then use student first job score: 0-2 = low, 3-4 = medium, 5-6 = high

* Income
  * Household income: 1-3 = low, 4-6 = medium, 7-8 = high

* combined score = total education + employment + income/3: 1-2 = low, 3-4 = medium, 5+ = high

# Compare SES levels against anxiety, stress and depression

See also: [one-way ANOVA](anova1.html#anovaWS)

```{r anova-anxiety, message=FALSE}
library(BayesFactor, quietly = TRUE)
# FIXME: can't use subj as random factor as only 1 observation per subj
# Error in base::try(expression, silent = silent) : not enough observations
bf <- anovaBF(formula = total ~ ses, data = anxiety)
bf
```

```{r anova-depression}
bf <- anovaBF(formula = total ~ ses, data = depression)
bf
```

```{r anova-stress}
bf <- anovaBF(formula = total ~ ses, data = stress)
bf
```

# SES category descriptives

See also: [better tables](better-tables.html)

```{r category-descriptives}
anxiety %>%
  group_by(ses_category) %>%
  summarise(mean = mean(total), sd = sd(total))

depression %>%
  group_by(ses_category) %>%
  summarise(mean = mean(total), sd = sd(total))

stress %>%
  group_by(ses_category) %>%
  summarise(mean = mean(total), sd = sd(total))
```

# Compare SES categories against anxiety, stress and depression

See also: [one-way ANOVA](anova1.html#anovaWS)

```{r anova-anxiety-category}
bf <- anovaBF(formula = total ~ ses_category, data = anxiety)
bf
```
```{r anova-depression-category}
bf <- anovaBF(formula = total ~ ses_category, data = depression)
bf
```

```{r anova-stress-category}
bf <- anovaBF(formula = total ~ ses_category, data = stress)
bf
```

# Pearson correlations

See also: [correlation](corr.html#corr)

* depression ~ anxiety
* depression ~ stress
* anxiety ~ stress

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

