---
title: "Cross-cultural relationships between trait emotional intelligence, religion and spirituality"
author: "Alison Bacon, Dawid Poblocki, Paul Sharpe and Andy Wills"
output:
  html_document:
    highlight: pygment
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate
## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)
## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
library(pander)
```

## Before you start

This is an advanced worksheet, which assumes you have completed the [Absolute Beginners' Guide to R](https://ajwills72.github.io/rminr/#beginners) course, the [Research Methods in Practice (Quantitative section)](https://benwhalley.github.io/rmip/overview-quantitative.html) course, and the [Intermediate Guide to R](https://ajwills72.github.io/rminr/#rmip) course. 

## Contents

- [Introduction](#intro)
- [Loading data](#load)

<a name="intro"></a>

## Introduction

This worksheet describes a full analysis pipeline for an undergraduate student dissertation which explored relationships between Polish (n=93) and British (n=104) people's emotional intelligence, religious orientation and spirituality.

In this study, emotional intelligence (EI) was treated as a trait; a personality factor relating to various aspects of emotions. Religious orientation (RO) was treated as either intrinsic or extrinsic. People who are intrinsically religious treat religion as a spiritual end in and of itself. Those who are extrinsically religious practise religion for self-serving reasons, such as social status. Spirituality was measured in terms of connectedness and its link to happiness. 

The researchers sought to replicate previous research which found that intrinsically religious people were more spiritual than extrinsically religiously people. They also predicted a negative correlation between EI and extrinsic RO. Finally, they predicted that intrinsic RO would correlate with trait EI.

<a name="load"></a>

## Loading data

Open the `rminr-data` project we used [previously](preproc.html#load).

Ensure you have the latest files by asking git to "`pull`" the repository. Select the `Git` tab, which is located in the row of tabs which includes the `Environment` tab. Click the `Pull` button with a downward pointing arrow. A window will open showing the files which have been pulled from the repository. Close the `Git pull` window. The `case-studies` folder should contain the folder `alison-bacon`.

Next, create a new, empty R script and save it in the `rminr-data` folder as `cs-alison-bacon.R`. Put all the commands from this worksheet into this file, and run them from there. Save your script regularly.

We start by loading the data.

```{r data-note, echo=FALSE, eval=FALSE}
# All data _except* TEI came from Alison, as she needed to adjust education and religion levels.
# TEI data came from Dawid, as Alison's data did not exclude subscale scores.
```

```{r excel, message=FALSE}
rm(list = ls()) # clear the environment
library(tidyverse)

# read data
raw_data <- read_csv('case-studies/alison-bacon/raw_data.csv')
```

197 participants (93 polish, 104 british) were analysed (exclusions: 10 missing data, 2 no consent) Polish/English.

### Religions orientation

See also: [preprocessing scales](preproc-scales.html), [analysing scales](analyse-scales.html).

Religious orientation (RO) was treated as either intrinsic or extrinsic. People who are intrinsically religious treat religion as a spiritual end in and of itself. Those who are extrinsically religious practise religion for self-serving reasons, such as social status. 
 
The amended Religious Orientation Scale (ROS; Maltby & Lewis, 1996), is a 20-item scale, with items scored from 1-3.

```{r ro}
# score Religious Orientation Scale

## FIXME: I: 25, 44, 90 = NA, also NA in O

# "we would have replaced [NA] with the mean, or with a mid-scale score"

ro_i <- c('RO1','RO5','RO6','RO9','RO11','RO12','RO16','RO19')
ro_e <- c('RO2','RO3','RO4','RO7','RO8','RO10','RO13','RO14','RO15','RO17','RO18','RO20')
data <- raw_data %>%
  rowwise() %>%
  mutate(ro_i = mean(c_across(all_of(ro_i)), na.rm = TRUE),
         ro_e = mean(c_across(all_of(ro_e))), na.rm = TRUE)
```

**Explanation of commands:**

### Spirituality

Spirituality was measured in terms of connectedness and its link to happiness. 

The Spiritual Connection Questionnaire (SCQ-48; Wheeler and Hyland, 2008) is a 48-item scale, with items scored on a seven point Likert scale. It has five factors: connection with nature, connection with places, connection with the universe, connection with other people, and the extent to which spirituality brings one happiness. 

The total score could be obtained by summing up scores for all the items, after reversing score for negative items (used to avoid acquiescence bias), which comprised half of the statements (24 items)."

```{r scq48}
# score SCQ-48
nature    <- c('scq32','scq8','scq17')
places    <- c('scq10','scq5','scq3','scq36','scq28')
universe  <- c('scq47','scq45','scq22','scq33','scq21','scq37','scq25',
               'scq42','scq29','scq4','scq16','scq46','scq38','scq18',
               'scq14','scq13','scq9','scq26','scq40')
others    <- c('scq41','scq11','scq48','scq19','scq15','scq30','scq34',
               'scq27','scq1','scq6','scq44','scq23')
happiness <- c('scq43','scq35','scq39','scq31','scq7','scq12','scq20',
               'scq24','scq2')

data <- data %>%
  rowwise() %>%
  mutate(happiness = mean(c_across(all_of(happiness)), na.rm = TRUE),
         places    = mean(c_across(all_of(places)), na.rm = TRUE),
         others    = mean(c_across(all_of(others)), na.rm = TRUE),
         nature    = mean(c_across(all_of(nature)), na.rm = TRUE),
         universe  = mean(c_across(all_of(universe)), na.rm = TRUE))
```

### Trait Emotional Intelligence

In this study, emotional intelligence (EI) was treated as a trait; a personality factor relating to various aspects of emotions.

The Short Form Trait Emotional Intelligence Questionnaire (TEIQue-SF, Petrides & Furnham, 2006), is a 30-item questionnaire, with items scored from 1-7. It consists of four factors: well-being, self-control, emotionality and sociability.

```{r tei}
# score TEIQue-SF
data <- data %>%
  select(subj, age, gender, education, religion,
         ro_i, ro_e, happiness, places, others, nature, universe,
         tei, wellbeing, self_control, emotionality, sociability)
```

## Create 'religious' binary factor

```{r}
data <- data %>% mutate(religious = ifelse(religion == 'other or none', 'yes', 'no')) %>%
  select(subj, age, gender, education, religious,
         ro_i, ro_e, happiness, places, others, nature, universe,
         tei, wellbeing, self_control, emotionality, sociability)
missing <- data %>% filter_all(any_vars(is.na(.)))
```

### Demographics

## Descriptives

See also [better tables](better-tables.html#descriptives).

"Table of descriptives (mean and SD) for the two groups separately (all SCQ subscales) and t tests to compare the two groups on all measures. They will differ significantly only on the two RO measures with Polish group higher scores in both cases."

## Correlation matrices (Tables 1 & 2)

See also: [better tables](better-tables.html#cor-matrix)

It should possible to define this as a comparison between two models.

"Correlations for both groups separately and observe differences"
      
* Polish: Religious Belief, Intrinsic RO, Extrinsic RO, EI, Spirituality
* British: Religious Belief, Intrinsic RO, Extrinsic RO, EI, Spirituality

##  Correlations between spirituality subdomains and religious orientation

See also: [better tables](better-tables.html#cor-matrix)

* Custom table (Table 3)

## Regression

See also: [more on regression](more-on-regression.html).

"Regression on TEI total score. If we are not covering moderation, they can do the two groups separately as they will see some differences on the SCQ subscales. Polish H and O are significant, UK H and P are significant. There are cultural differences in things like attitudes to extended family and collectivism which could explain this."

## References

Maltby, J., & Lewis, C. A. (1996). Measuring intrinsic and extrinsic orientation toward
religion: Amendments for its use among religious and non-religious samples.
Personality and Individual Differences, 21(6), 937-946.

Petrides, K. V., & Furnham, A. (2006). The Role of Trait Emotional Intelligence in a
Gender‐Specific Model of Organizational Variables 1. Journal of Applied
Social Psychology, 36(2), 552-569.

Wheeler, P., & Hyland, M. E. (2008). The development of a scale to measure the
experience of spiritual connection and the correlation between this experience
and values. Spirituality and Health International, 9(4), 193-217.

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

