---
title: "Links between openness to experience, imagery and creative problem solving"
author: "Jon May, Jessica Hart, Paul Sharpe and Andy Wills"
output:
  html_document:
    highlight: pygment
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate
## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)
## See also
## https://stackoverflow.com/questions/39709745/decreasing-the-line-thickness-and-corr-font-size-in-ggpairs-plot
## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
library(pander)
```

## Before you start

This is an advanced worksheet, which assumes you have completed the [Absolute Beginners' Guide to R](https://ajwills72.github.io/rminr/#beginners) course, the [Research Methods in Practice (Quantitative section)](https://benwhalley.github.io/rmip/overview-quantitative.html) course, and the [Intermediate Guide to R](https://ajwills72.github.io/rminr/#rmip) course. 

## Contents

- [Introduction](#intro)
- [Loading data](#load)
- [Correlation between openness and problems solved](#cor-o-problems)
- [Correlation matrix: other personality factors and problems solved](#cor-matrix)

<a name="intro"></a>

## Introduction

This worksheet describes a full analysis pipeline for an undergraduate student dissertation which explored relationships between personality, imagery and creative problem solving. Forty-eight students were tested to address three hypotheses. First, the researchers predicted that participants with more open personality types would be better at solving a selection of problems requiring creative solutions. Second, they predicated that participants with more vivid mental imagery would be better at solving the problems. Third, they predicted a relationship between divergent thinking and an ability to solve the problems.

Personality was measured using a simplified version of Costa and McCrae’s (1992) "big five" personality questionnaire. Mental imagery was measured using the PsiQ Plymouth Sensory Imagery Questionnaire (Andrade et al., 2013). Divergent thinking was measured using a 'flexible thinking task', which measured fluency, flexibility and originality. The problems requiring creative solutions were taken from May (1987).

<a name="load"></a>

## Loading data

Open the `rminr-data` project we used [previously](preproc.html#load).

Ensure you have the latest files by asking git to "`pull`" the repository. Select the `Git` tab, which is located in the row of tabs which includes the `Environment` tab. Click the `Pull` button with a downward pointing arrow. A window will open showing the files which have been pulled from the repository. Close the `Git pull` window. The `case-studies` folder should contain the folder `jon-may`.

Next, create a new, empty R script and save it in the `rminr-data` folder as `cs-jon-may.R`. Put all the commands from this worksheet into this file, and run them from there. Save your script regularly.

We start by loading the data.

```{r excel, message=FALSE}
rm(list = ls()) # clear the environment
library(tidyverse)
```

**Explanation of commands:**

We clear the workspace, and load `tidyverse` package, then read the four data files.

<a name="preproc"></a>

## Preprocessing

### Problems

Problem booklet with 7 problems (solved=1, not=0, total=n solved)

```{r}
problems <- read_csv('case-studies/jon-may/problems.csv')
solved <- problems %>%
  rowwise() %>%
  mutate(problems = sum(c_across(bridges:dots))) %>%
  select(subj, problems)
```

### OCEAN

Big5 OCEAN questionnaire 50 items answered 1 to 5 (includes a data entry error)

```{r}
ocean   <- read_csv('case-studies/jon-may/big5.csv')
#big5    <- read_csv('case-studies/jon-may/big5_total.csv', col_types = 'fiiiii')

oceankey<-read.csv("case-studies/jon-may/oceankey.csv")

scaleIDs <- oceankey$ScaleID                 # make a vector of the new variable names
names(scaleIDs) <- oceankey$ItemID           # name each new name with the old name so it can be looked-up

ocean.scales<-ocean %>%
   pivot_longer(S1:S50,names_to='Item', values_to='Value') %>%     # put the 50 items into a column
   mutate(ScaleID=scaleIDs[Item])%>%                               # lookup the new names from the vector
   select(-Item) %>%                                               # drop the columns we don't need
   mutate(Value=ifelse(Value>5,NA,ifelse(Value<1, NA, Value))) %>% # screen for absurd values
   mutate(Value=ifelse(grepl("r",ScaleID), 6-Value, Value)) %>%    # reverse score items containg 'r' ...
   mutate(ScaleID = sub("r", "", ScaleID)) %>%                     # ... and remove the 'r'
   pivot_wider(names_from="ScaleID", values_from="Value") %>%      # make the data wide again
   select(subj,sort(names(.))) %>%                                 # sort the columns by the new names
   rowwise() %>%                                                   # for each participant...
   mutate(openness=mean(c_across(o01:o10), na.rm = TRUE),          # ... mean of 10 items in each scale
          conscientiousness=mean(c_across(C01:C10), na.rm = TRUE),
          extraversion=mean(c_across(E01:E10), na.rm = TRUE),
          agreeableness=mean(c_across(A01:A10), na.rm = TRUE),
          neuroticism=mean(c_across(N01:N10), na.rm = TRUE)) %>%
   select(subj, openness, conscientiousness, extraversion, agreeableness, neuroticism)
```

### Flexible thinking test

* We don't have raw data for the ‘Flexible Thinking Test’ (overall scores mean of ftt1-3)

```{r}
ftt <- read_csv('case-studies/jon-may/ftt.csv') %>%
   rowwise() %>%
   mutate(ftt = mean(c_across(ftt1:ftt3))) %>%
   select(subj, ftt)
```

* the 35 PsiQ vividness of imagery items - 5 of each of 7 modalities (0 not at all to 10 as vivid as real life) (score = mean of all items)

```{r}
psiq <- read_csv('case-studies/jon-may/psiq.csv') %>%
   rowwise() %>%
   mutate(psiq = mean(c_across(2:36))) %>%
   select(subj, psiq)
```

```{r}
data <- full_join(ocean.scales, solved, by = 'subj') %>%
   full_join(ftt, by = 'subj') %>%
   full_join(psiq, by = 'subj')
```

## Mean, Standard deviation, skewness, kurtosis for all independent variables (n=48)

The `psych` library has a useful function `describe` to obtain descriptive statistics.

```{r}
library(psych)

describe(data)
```

## Histogram and curve showing distribution of problem scores (n=48)

https://stackoverflow.com/questions/6967664/ggplot2-histogram-with-normal-curve

```{r}
n_obs = sum(!is.na(data$problems))
bw = 1
mean <- mean(data$problems)
sd <- sd(data$problems)

data %>% ggplot(aes(problems))  + 
  geom_histogram(colour = "black", binwidth = bw) + 
  stat_function(fun = function(x) 
    dnorm(x, mean = mean, sd = sd) * bw * n_obs) +
  xlab('Problems solved') + ylab('Count (participants)')
```

## Scatterplot and best fit line, openness vs. problems solved

You learnt how to create a scatterplot and a best fit line in the [regression worksheet](https://benwhalley.github.io/rmip/regression.html).

<a name="cor-o-problems"></a>

```{r}
data %>%
  ggplot(aes(openness, problems)) +
  geom_point() +
  geom_smooth(method=lm, se=F)

```

## Correlation between openness and problems solved

The researchers predicted that participants with more open personality types would be better at solving the problems. In other words, we should expect a positive correlation between openness and problem solving. We test a directional hypothesis with a one-tailed correlation. See the [More on relationships, part 2.](#corr-extended.html) worksheet for more details.

```{r}
cor_o_problems<-cor.test(data$problems, data$openness, alternative='greater')
cor_o_problems

# note that this is a 2-tailed test
library(BayesFactor)
cor_o_problems_bf<-correlationBF(data$problems, data$openness)
cor_o_problems_bf
```

**Explanation of commands:**

We use a Pearson correlation to look at the relationship between openness and the number of problems solved. We specify `alternative='greater'` to indicate that we are predicting a positive correlation (one-tailed test). We also calculate a Bayes factor to test the evidence for the correlation.

**Explanation of output:**

Contrary to the first hypothesis, there is no evidence for a positive correlation  between openness and creative problem solving  (_r_ = `r round(cor_o_problems$estimate,2)`, one-tailed, _BF_ = `r round(BayesFactor::extractBF(cor_o_problems_bf)$bf,2)`).

<a name="cor-matrix"></a>

## Correlation matrix: other personality factors and problems solved

Correlation matrices are covered in the [Better tables](better-tables.html#cor-matrix) worksheet.

```{r cor-matrix, R.options = list(width = 10000)}
library(apaTables)
apa.cor.table(data %>% select(conscientiousness:problems), filename='table1.doc', table.number = 1)
```

**Explanation of output:**

There were no correlations between the other four personality factors and problem solving. Note that these are 2-tailed correlations, as it's not possible to specify 1-tailed tests with `apa.cor.table()`.


## Correlation between problems solved and flexible thinking

```{r}
cor_f_problems<-cor.test(data$problems, data$ftt, alternative='greater')
cor_f_problems

# note that this is a 2-tailed test
library(BayesFactor)
cor_f_problems_bf<-correlationBF(data$problems, data$ftt)
cor_f_problems_bf
```
## Correlation between problems solved and vividness

```{r}
cor_v_problems<-cor.test(data$problems, data$psiq, alternative='greater')
cor_v_problems

# note that this is a 2-tailed test
library(BayesFactor)
cor_v_problems_bf<-correlationBF(data$problems, data$psiq)
cor_v_problems_bf
```
**Explanation of commands:**

As for Openness, we use Pearson correlations to look at the relationship between the number of problems solved, Flexible Thinking and Vividness. We specify `alternative='greater'` to indicate that we are predicting a positive correlation (one-tailed test). We also calculate a Bayes factor to test the evidence for the correlations.

**Explanation of output:**

There is evidence for a positive correlation between flexibility and creative problem solving  (_r_ = `r round(cor_f_problems$estimate,2)`, one-tailed, _BF_ = `r round(BayesFactor::extractBF(cor_f_problems_bf)$bf,2)`). There is no evidence for a positive correlation between imagery vividness and creative problem solving  (_r_ = `r round(cor_v_problems$estimate,2)`, one-tailed, _BF_ = `r round(BayesFactor::extractBF(cor_v_problems_bf)$bf,2)`).


## Pairs plot

Pairs plots are are covered in the [Better graphs](better-graphs.html#pairs) worksheet.

```{r pairs, class.source = 'numberLines lineAnchors', message=FALSE}
library(GGally)
#source('themeapa.R')
data %>%
  select(psiq, openness, problems, ftt) %>%
  ggpairs(lower=list(continuous='smooth')) #+
  #theme_APA
```

## References

Andrade, J., May, J., Deeprose, C., Baugh, S.-J., & Ganis, G. (2014). [Assessing vividness of mental imagery: The Plymouth Sensory Imagery Questionnaire](https://onlinelibrary.wiley.com/doi/pdf/10.1111/bjop.12050?casa_token=-IMvMthFsKQAAAAA:_b6s7nwZ8JS-5fWM4Xxgey5sBBPpbIRPUupgZ2XFQ-jNYGNOK0ARomtRNgCIWUfZqUD7w9PBoLbeng) British Journal of Psychology, 105(4), 547–563. 

Costa, P. T., & McCrae, R. R. (1992). Neo personality inventory-revised (NEO PI-R). 
Psychological Assessment Resources Odessa, FL.

May, J. (1987). The cognitive analysis of flexible thinking. Unpublished PhD
thesis, University of Exeter.

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

