---
title: "The Effect of Alternative Music on Emotional Response in Alternative Music Subcultures"
author: "Helen Lloyd, Elizabeth Dark, Paul Sharpe and Andy Wills"
output:
  html_document:
    highlight: pygment
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate
## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
library(pander)
```

## Contents

- [Introduction](#intro)
- [Loading data](#load)
- [Excluding data by subject number](#exclude)
- [Preprocessing](#preprocess)
- [Compare avoidance, approach and self-development strategies by subculture](#ers-subscales)

## Before you start

This is an advanced worksheet, which assumes you have completed the [Absolute Beginners' Guide to R](https://ajwills72.github.io/rminr/#beginners) course, the [Research Methods in Practice (Quantitative section)](https://benwhalley.github.io/rmip/overview-quantitative.html) course, and the [Intermediate Guide to R](https://ajwills72.github.io/rminr/#rmip) course. 

<a name="intro"></a>

## Introduction

This worksheet describes a full analysis pipeline for an undergraduate student dissertation which compared emotional responses before and after listening to different types of music. Participants self-identified as fans of emo (81), goth (77), metal (77) or mainstream (59) music. Their emotions were tested using the short-form Positive and Negative Affect Scale (PANAS-SF, Watson et al., 1988), before and after listening to 4-5 minutes of music associated with their subculture. The 'mainstream' group acted as a control condition. After the second PANAS-SF, they completed the Emotion Regulation Strategies for Artistic Creative Activities Scale (ERS-ACA, Fancourt et al., 2019). This was an online study, with data collected using [JISC](https://www.jisc.ac.uk/).

So there were two dependent variables in this study. The PANAS-SF was a dependent variable in a mixed design, with subculture group as the between subjects factor, and time (before and after the music) as the repeated measure. The other dependent variable, ERS-ACA, was analysed in relation to the subculture group.

"Hypothesis 1: whether subculture-selected music
elicited emotional response in each group to the same extent, and in comparison to
the Mainstream group before and after music"

"Hypothesis 2: ERS ACA to investigate whether respondents use music to regulate emotion in different ways between groups."

<a name="load"></a>

## Loading data

Open the `rminr-data` project we used [previously](preproc.html#load).

Ensure you have the latest files by asking git to "`pull`" the repository. Select the `Git` tab, which is located in the row of tabs which includes the `Environment` tab. Click the `Pull` button with a downward pointing arrow. A window will open showing the files which have been pulled from the repository. Close the `Git pull` window. The `case-studies` folder should contain a folder named `helen-lloyd`.

Next, create a new, empty R script and save it in the `rminr-data` folder as `cs-music-emotion.R`. Put all the commands from this worksheet into this file, and run them from there. Save your script regularly.

We start by loading the data.

```{r load, message=FALSE, class.source = 'numberLines lineAnchors'}
rm(list = ls()) # clear the environment
library(tidyverse)
raw <- read_csv('case-studies/helen-lloyd/music-emotion.csv')
```

**Explanation of commands:**

Line 1 clears the environment. Line 2 loads `tidyverse`. Line 3 loads the data into the `raw` data frame.

<a name="exclude"></a>

## Excluding data by subject number

See also: [excluding participants](preproc-experiments.html#exclude)

In this study, the researcher excluded participants they considered to be extreme outliers, and those with incomplete data.

```{r exclude, class.source = 'numberLines lineAnchors'}
## Exclude participants
exclude <- c(24, 158, 234, 405, 497, 630, 663, 673, 685)
include <- raw %>% filter(!(subj %in% exclude))
```

### Explanation of commands

Excluding participants by subject number is covered in the [Data preprocessing for experiments worksheet](#preproc-experiments.html). In summary, line 1 is a list of the participant numbers that we want to exclude. Line 2 excludes these participants by filtering `raw` to include values in `sub` which are _not_ in `exclude`.

## Preprocessing

<a name="preprocess"></a>

```{r tidy, message=FALSE}
include <- include %>%
  set_names(~ str_to_lower(.)) %>%
  mutate(subj = factor(subj), subculture = factor(subculture)) %>%
  mutate(subculture = recode(.$subculture, Main = 'Mainstream')) %>%
  select(subj, subculture, interestedpre:ersq18)
include %>% head(3) %>% pander(split.table = Inf)
```

## Score PANAS-SF

See also: [preprocessing scales](preproc-scales.html), [analysing scales](analyse-scales.html).

```{r score-panas}
panas <- include %>%
  mutate(pre_emotion = rowSums(across(interestedpre:afraidpre)),
         pre_pa = rowSums(across(c(interestedpre, excitedpre, strongpre,
                                   enthusiasticpre, proudpre, alertpre,
                                   inspiredpre, determinedpre, attentivepre,
                                   activepre))),
         pre_na = rowSums(across(c(distressedpre, upsetpre, guiltypre,
                                   scaredpre, hostilepre, irritablepre,
                                   ashamedpre, nervouspre, jitterypre,
                                   afraidpre))),
         post_emotion = rowSums(across(interestedpost:afraidpost)),
         post_pa = rowSums(across(c(interestedpost, excitedpost, strongpost,
                                    enthusiasticpost, proudpost, alertpost,
                                    inspiredpost, determinedpost,
                                    attentivepost, activepost))),
         post_na = rowSums(across(c(distressedpost, upsetpost, guiltypost,
                                    scaredpost, hostilepost, irritablepost,
                                    ashamedpost, nervouspost, jitterypost,
                                    afraidpost))),
         ) %>%
  select(subj, subculture, pre_emotion:post_na)
panas %>% head(3) %>% pander(split.table = Inf)
```

## Score ERS-ACA

The Emotion Regulation Strategies for Artistic Creative Activities Scale (ERS-ACA; Fancourt, Garnett, Spiro, West, Mullensiefen 2019) is an 18 item inventory, with each item scored from 1 ('strongly disagree') to 5 ('strongly agree'). The ERS-ACA gives an overall measure of the strategy people use to regulate their emotions when they engage in artistic, creative activities, and scores on three strategy sub-scales; avoidance, approach and self-development.


```{r ers}
avoidance_q   <- c('ersq1', 'ersq3', 'ersq4', 'ersq8', 'ersq11', 'ersq14', 'ersq18')
approach_q    <- c('ersq2', 'ersq9', 'ersq10', 'ersq12', 'ersq13', 'ersq16')
development_q <- c('ersq5', 'ersq6', 'ersq7', 'ersq15', 'ersq17')
ers <- include %>%
  mutate(avoidance = rowSums(across(all_of(avoidance_q))) / length(avoidance_q),
         approach = rowSums(across(all_of(approach_q))) / length(approach_q),
         development = rowSums(across(all_of(development_q))) / length(development_q),
         total = rowSums(across(all_of(c(avoidance_q, approach_q, development_q))) / 18)) %>%
  select(subj, subculture, avoidance, approach, development, total)
```

### Explanation of commands

## TODO: Tables: for each emotion,PA,NA: mean(sd), time*subculture (Appendix 2)

See also: [better tables](better-tables.html)

## TODO: 3 ANOVA plots: : PANAS ~ subculture(4)*time(2) (Appendices 6-8)

See also: [better graphs](better-graphs.html)

QUESTION: How are "estimated marginal means" calculated?

## TODO: 3 mixed ANOVAs: PANAS ~ subculture(4)*time(2)

See also: [factorial ANOVA](anova3.html)

### Reciprocal transform NA

See also: [reciprocal transform](preproc-experiments.html#ex1)

## ERS summary statistics

```{r ers-summary}
ers_l <- ers %>%
  pivot_longer(avoidance:total, names_to = 'ers', values_to = 'score') %>%
  mutate(ers = factor(ers))
ers_l %>%
  group_by(ers, subculture) %>%
  summarise(mean = mean(score, na.rm = TRUE), sd = sd(score, na.rm = TRUE))
```

## Compare overall emotional regulation strategy by subculture

```{r ers-anova}
library(BayesFactor)
ers_total <- ers_l %>% filter(ers == 'total') %>% drop_na()
anovaBF(formula = score ~ subculture, data = ers_total)
```

### Explanation of commands

### Explanation of output

<a name="ers-subscales"></a>

## Compare avoidance, approach and self-development strategies by subculture

### Avoidance

The Kruskal-Wallis _H_ test is described in the [Traditional non-parametric tests worksheet](non-parametric.html).

```{r avoidance}
library("ggpubr")
avoidance <- ers_l %>% filter(ers == 'avoidance')
ggboxplot(avoidance, x = 'subculture', y = 'score', 
          color = 'subculture', order = c('Mainstream', 'Goth', 'Emo', 'Metal'),
          ylab = "ERS avoidance", xlab = "Subculture")
avoidance <- avoidance %>% mutate(rank = rank(score))
avoidance %>%
  group_by(subculture) %>%
  summarise(n = n(),
            median = median(score, na.rm = TRUE),
            mean_rank = mean(rank),
            sum_rank = sum(rank))
kruskal.test(score ~ subculture, data = avoidance)
```

### Approach

```{r approach}
approach <- ers_l %>% filter(ers == 'approach')
approach <- approach %>% mutate(rank = rank(score))
approach %>%
  group_by(subculture) %>%
  summarise(n = n(),
            median = median(score, na.rm = TRUE),
            mean_rank = mean(rank),
            sum_rank = sum(rank))
ggboxplot(approach, x = 'subculture', y = 'score', 
          color = 'subculture', order = c('Mainstream', 'Goth', 'Emo', 'Metal'),
          ylab = "ERS approach", xlab = "Subculture")
kruskal.test(score ~ subculture, data = approach)
```

## Self-development

```{r development}
development <- ers_l %>% filter(ers == 'development')
development <- development %>% mutate(rank = rank(score))
development %>%
  group_by(subculture) %>%
  summarise(n = n(),
            median = median(score, na.rm = TRUE),
            mean_rank = mean(rank),
            sum_rank = sum(rank))
ggboxplot(development, x = 'subculture', y = 'score', 
          color = 'subculture', order = c('Mainstream', 'Goth', 'Emo', 'Metal'),
          ylab = "ERS self-development", xlab = "Subculture")
kruskal.test(score ~ subculture, data = development)
```

## References

Watson, D., Clark, L. A., & Tellegen, A. (1988). [Development and validation of brief measures of positive and negative affect: The PANAS scales.](https://www.researchgate.net/profile/Rachel_Bachner-Melman/post/Is_there_a_way_to_measure_negative_emotions/attachment/59d644a1c49f478072ead59a/AS:273817966972930@1442294698972/download/PANAS.+20+items.pdf) Journal of Personality and Social Psychology, 54(6), 1063.

Fancourt, D., Garnett, C., Spiro, N., West, R., & Müllensiefen, D. (2019). [How do artistic creative activities regulate our emotions? Validation of the Emotion Regulation Strategies for Artistic Creative Activities Scale (ERS-ACA)](https://doi.org/10.1371/journal.pone.0211362). PLOS ONE, 14(2), e0211362. 

___
  
  This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

