---
title: "The Effect of Alternative Music on Emotional Response in Alternative Music Subcultures"
author: "Helen Lloyd, Elizabeth Dark, Paul Sharpe"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Data required to knit
## https://github.com/ajwills72/rminr-data/tree/master/going-further/music-emotion.csv
##
## I check out rminr-data and make a symbolic link to going-further

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
options(tibble.width = Inf) # show all columns in output
library(pander)
```

# Contents

- [Introduction](#intro)

- [Getting started](#start)

- [Preprocessing](#preprocessing)

<a name="intro"></a>

# Introduction

* PANAS-SF (20 item) pre and post
* ERS ACA administered after post-PANAS-SF

## TODO: Exclusion criteria

<a name="start"></a>

# Getting started

```{r init-load, message=FALSE}
rm(list = ls()) # clear the environment
library(tidyverse)
```

<a name="preprocessing"></a>

# Preprocessing

See also: [Preprocessing scales](#preproc-scales.html)

```{r tidy, message=FALSE}
data1 <- read_csv('going-further/music-emotion-1.csv')

d1 <- data1 %>%
  set_names(~ str_to_lower(.)) %>%
  mutate(subj = factor(partexpid), subculture = factor(subcultureauto)) %>%
  select(subj, interestedpre:afraidpost, subculture)
d1 %>% head(3) %>% pander(split.table = Inf)
```

```{r score}
d1 <- d1 %>% mutate(pre_emotion = rowSums(across(interestedpre:afraidpre)),
                    pre_pa = rowSums(across(c(interestedpre, excitedpre, strongpre, enthusiasticpre,
                                            proudpre, alertpre, inspiredpre, determinedpre, attentivepre,
                                            activepre))),
                    pre_na = rowSums(across(c(distressedpre, upsetpre, guiltypre, scaredpre,
                                                  hostilepre, irritablepre, ashamedpre, nervouspre,
                                                  jitterypre, afraidpre))),
                    post_emotion = rowSums(across(interestedpost:afraidpost)),
                    post_pa = rowSums(across(c(interestedpost, excitedpost, strongpost, enthusiasticpost,
                                            proudpost, alertpost, inspiredpost, determinedpost, attentivepost,
                                            activepost))),
                    post_na = rowSums(across(c(distressedpost, upsetpost, guiltypost, scaredpost,
                                                  hostilepost, irritablepost, ashamedpost, nervouspost,
                                                  jitterypost, afraidpost))),
                    ) %>%
  select(subj, subculture, pre_emotion:post_na)
d1 %>% head(3) %>% pander(split.table = Inf)
```

QUESTION: these scores matched those in the data, so presumably I scored the PANAS-SF correctly?

## ERS scoring

QUESTION: Should we simplify things by using data1.csv for the ERS analyses?


"The Emotion Regulation Strategies for Artistic Creative Activities Scale (ERS ACA)
(Fancourt, Garnett, Spiro, West, Mullensiefen 2019) is an 18 item inventory with a 5
point scale from ‘strongly disagree’ to ‘strongly agree’. This was designed to
measure the extent to which a person uses regulation strategies while engaging with
an art activity. An overall general strategy use is calculated, as well as 3 subscales
of strategies: Avoidance, Approach, and Self Development Strategies."

QUESTION: Score the ERS as per Fancourt et al. (2019, Table 3)? No reversed items?

```{r data2}
data2 <- read_csv('going-further/music-emotion-2.csv')
```

# Results

## TODO: Tables: for each emotion,PA,NA: mean(sd), time*subculture (Appendix 2)

## TODO: 3 ANOVA plots: : PANAS ~ subculture(4)*time(2) (Appendices 6-8)

## TODO: 3 mixed ANOVAs: PANAS ~ subculture(4)*time(2)

## TODO: 1-way ANOVAs: overal ERS ~ subculture

## TODO: 3 Kruskal-Wallis H tests: ERS_subscale ~ subculture

# References

Watson, D., Clark, L. A., & Tellegen, A. (1988). Development and validation of brief measures of positive and negative affect: The PANAS scales. Journal of Personality and Social Psychology, 54(6), 1063.

Fancourt, D., Garnett, C., Spiro, N., West, R., & Müllensiefen, D. (2019). How do artistic creative activities regulate our emotions? Validation of the Emotion Regulation Strategies for Artistic Creative Activities Scale (ERS-ACA). PLOS ONE, 14(2), e0211362. https://doi.org/10.1371/journal.pone.0211362

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

