---
title: "The Effect of Alternative Music on Emotional Response in Alternative Music Subcultures"
output:
  html_document:
    highlight: pygment
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Data required to knit
## https://github.com/ajwills72/rminr-data/tree/master/going-further/music-emotion.csv
##
## I check out rminr-data and make a symbolic link to going-further

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
library(pander)
```

## Contents

- [Introduction](#intro)
- [Loading data](#load)
- [Excluding data by subject number](#exclude)
- [Preprocessing](#preprocess)

## Before you start

This is an advanced worksheet, which assumes you have completed the [Absolute Beginners' Guide to R](https://ajwills72.github.io/rminr/#beginners) course, the [Research Methods in Practice (Quantitative section)](https://benwhalley.github.io/rmip/overview-quantitative.html) course, and the [Intermediate Guide to R](https://ajwills72.github.io/rminr/#rmip) course. 

<a name="intro"></a>

## Introduction

This worksheet describes a full analysis pipeline for an undergraduate student dissertation which compared emotional responses before an after listening to different types of music. Participants self-identified as fans of emo (81), goth (77), metal (77) or mainstream (59) music. Their emotions were tested using the short-form Positive and Negative Affect Scale (PANAS-SF, Watson et al., 1988), before and after listening to 4-5 minutes of music associated with their subculture. The 'mainstream' group acted as a control condition. After the second PANAS-SF, they completed the Emotion Regulation Strategies for Artistic Creative Activities Scale (ERS-ACA, Fancourt et al., 2019). This was an online study, with data collected using [JISC](https://www.jisc.ac.uk/).

So there were two dependent variables in this study. The PANAS-SF was a dependent variable in a mixed design, with subculture group as the between subjects factor, and time (before and after the music) as the repeated measure. The other dependent variable, ERS-ACA, was analysed in relation to the subculture group.

<a name="load"></a>

## Loading data

Open the `rminr-data` project we used [previously](preproc.html#load).

Ensure you have the latest files by asking git to "`pull`" the repository. Select the `Git` tab, which is located in the row of tabs which includes the `Environment` tab. Click the `Pull` button with a downward pointing arrow. A window will open showing the files which have been pulled from the repository. Close the `Git pull` window. The `case-studies` folder should contain a folder named `helen-lloyd`.

Next, create a new, empty R script and save it in the `rminr-data` folder as `cs-music-emotion.R`. Put all the commands from this worksheet into this file, and run them from there. Save your script regularly.

We start by loading the data.

```{r load, message=FALSE, class.source = 'numberLines lineAnchors'}
rm(list = ls()) # clear the environment
library(tidyverse)
raw <- read_csv('case-studies/helen-lloyd/music-emotion.csv')
```

**Explanation of commands:**

Line 1 clears the environment. Line 2 loads `tidyverse`. Line 3 loads the data into the `raw` data frame.

<a name="exclude"></a>

## Excluding data by subject number

In this study, the researcher excluded participants they considered to be extreme outliers, and those with incomplete data.

```{r exclude, class.source = 'numberLines lineAnchors'}
## Exclude participants
exclude <- c(24, 158, 234, 405, 497, 630, 663, 673, 685)
include <- raw %>% filter(!(subj %in% exclude))
```

### Explanation of commands

Excluding participants by subject number is covered in the [Data preprocessing for experiments worksheet](#preproc-experiments.html). In summary, line 1 is a list of the participant numbers that we want to exclude. Line 2 excludes these participants by filtering `raw` to include values in `sub` which are _not_ in `exclude`.

## Preprocessing

<a name="preprocess"></a>

```{r tidy, message=FALSE}
include <- include %>%
  set_names(~ str_to_lower(.)) %>%
  mutate(subj = factor(subj), subculture = factor(subculture)) %>%
  select(subj, interestedpre:afraidpost, subculture)
include %>% head(3) %>% pander(split.table = Inf)
```

## Score PANAS-SF

These scores matched those in the data, so presumably I scored the PANAS-SF correctly.

```{r score-panas}
panas <- include %>%
  mutate(pre_emotion = rowSums(across(interestedpre:afraidpre)),
         pre_pa = rowSums(across(c(interestedpre, excitedpre, strongpre,
                                   enthusiasticpre, proudpre, alertpre,
                                   inspiredpre, determinedpre, attentivepre,
                                   activepre))),
         pre_na = rowSums(across(c(distressedpre, upsetpre, guiltypre,
                                   scaredpre, hostilepre, irritablepre,
                                   ashamedpre, nervouspre, jitterypre,
                                   afraidpre))),
         post_emotion = rowSums(across(interestedpost:afraidpost)),
         post_pa = rowSums(across(c(interestedpost, excitedpost, strongpost,
                                    enthusiasticpost, proudpost, alertpost,
                                    inspiredpost, determinedpost,
                                    attentivepost, activepost))),
         post_na = rowSums(across(c(distressedpost, upsetpost, guiltypost,
                                    scaredpost, hostilepost, irritablepost,
                                    ashamedpost, nervouspost, jitterypost,
                                    afraidpost))),
         ) %>%
  select(subj, subculture, pre_emotion:post_na)
panas %>% head(3) %>% pander(split.table = Inf)
```

## Score ERS-ACA

"The Emotion Regulation Strategies for Artistic Creative Activities Scale (ERS ACA)
(Fancourt, Garnett, Spiro, West, Mullensiefen 2019) is an 18 item inventory with a 5
point scale from ‘strongly disagree’ to ‘strongly agree’. This was designed to
measure the extent to which a person uses regulation strategies while engaging with
an art activity. An overall general strategy use is calculated, as well as 3 subscales
of strategies: Avoidance, Approach, and Self Development Strategies."

TODO: Score the ERS as per Elizabeth's document and compare against Appendix 4

```{r ers}
```

## TODO: Tables: for each emotion,PA,NA: mean(sd), time*subculture (Appendix 2)

## TODO: 3 ANOVA plots: : PANAS ~ subculture(4)*time(2) (Appendices 6-8)

QUESTION: How are "estimated marginal means" calculated?

## TODO: 3 mixed ANOVAs: PANAS ~ subculture(4)*time(2)

## TODO: 1-way ANOVAs: overal ERS ~ subculture

## TODO: 3 Kruskal-Wallis H tests: ERS_subscale ~ subculture (4)

## References

Watson, D., Clark, L. A., & Tellegen, A. (1988). [Development and validation of brief measures of positive and negative affect: The PANAS scales.](https://www.researchgate.net/profile/Rachel_Bachner-Melman/post/Is_there_a_way_to_measure_negative_emotions/attachment/59d644a1c49f478072ead59a/AS:273817966972930@1442294698972/download/PANAS.+20+items.pdf) Journal of Personality and Social Psychology, 54(6), 1063.

Fancourt, D., Garnett, C., Spiro, N., West, R., & Müllensiefen, D. (2019). [How do artistic creative activities regulate our emotions? Validation of the Emotion Regulation Strategies for Artistic Creative Activities Scale (ERS-ACA)](https://doi.org/10.1371/journal.pone.0211362). PLOS ONE, 14(2), e0211362. 

  ___
  
  This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

