---
title: "The effects of negative mental imagery on self-esteem"
author: "Sophie Homer, Paul Sharpe"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Data required to knit
## https://github.com/ajwills72/rminr-data/tree/master/going-further/sses.csv
##
## I check out rminr-data and make a symbolic link to going-further

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
options(tibble.width = Inf) # show all columns in output
```

# Contents

- [Introduction](#intro)

- [Getting started](#start)

- [Preprocessing](#preprocessing)

- [Results](#results)

<a name="intro"></a>

# Introduction

This study was an experiment in which self-esteem was measured before and after a mental imagery intervention. Self-esteem was measured using the State Self-Esteem Scale (SSES, Osman et al., 2012), a 20-item survey used to measure short-lived (state) changes in self-esteem. Participants completed one of two mental imagery conditions, or a control condition:

 * Condition 1: participants visualised a negative mental image of themself
 * Condition 2: participants visualised a negative mental image of someone else
 * Condition 0 (control): participants did a card sorting task and did not think of any images.

This was a 2 (time) x 3 (condition) mixed design.

<a name="start"></a>

# Getting started

1. Open the project you used to complete the [Preprocessing scales](preproc-scales.html) worksheet.

1. You should have a script names `scales.R`. Add the code in this worksheet to the end of that script.

<a name="preprocessing"></a>

# Preprocessing

Run `scales.R` to preprocess the SSES data.

```{r preprocessing}
rm(list = ls()) # clear the environment
library(tidyverse)

sses <- read_csv('going-further/sses.csv')
sses_pre_raw  <- sses %>% select(1, 5:25) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition), time = factor('pre')) %>%
  select(subj, condition, time, q1:q20)
sses_post_raw <- sses %>% select(1, 25:45) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("post_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition), time = factor('post')) %>%
  select(subj, condition, time, q1:q20)
# Reverse score survey items
# @param scores raw scores data frame
# @param reverse character vector of columns to be reverse scored
# @param max integer value of maximum score
reverse_score_survey <- function(scores, reverse, max) {
  reverse_score <- function(score) { max - score } # for scores starting at 1, max + 1 - score
  scores %>% mutate_at(reverse, reverse_score)
}
reverse_questions <- c('q2', 'q4', 'q5', 'q7', 'q8', 'q10', 'q13', 'q15', 'q16', 'q17', 'q18',
                       'q19', 'q20')
sses_pre  <- reverse_score_survey(sses_pre_raw, reverse_questions, 4)
sses_post <- reverse_score_survey(sses_post_raw, reverse_questions, 4)
sses_pre  <- mutate(sses_pre, total = rowSums(sses_pre[4:23]))
sses_post <- mutate(sses_post, total = rowSums(sses_post[4:23]))
```

If you've done this correctly, `sses_pre` will look like this:

```{r}
library(pander)
sses_pre %>% head(3) %>% pander(split.table = Inf)
```

## Recode factor levels

Looking at the `condition` column above, we can see that the experimental software that generated our data used numbers to represent different levels of this factor. It's often clearer if the numbers are converted to strings which describe the factor levels.

```{r}
sses_pre$condition <- recode_factor(sses_pre$condition, `0` = 'control',
                                        `1` = 'self', `2` = 'other')
sses_post$condition <- recode_factor(sses_post$condition, `0` = 'control',
                                        `1` = 'self', `2` = 'other')
```

**Explanation of command:**

1. We use `recode_factor()` to rename the levels of the experimental condition (`sses_pre_raw$condition`). The current factor level name goes on the left hand side of the `=`, the new name goes on the right hand side. Notice that the numbers need to be in backticks. The renamed factor levels are reassigned to the same column `sses_pre_raw$condition`.

Our `condition` factor now has intuitive level names:

```{r}
sses_pre %>% head(3) %>% pander(split.table = Inf)
```

<a name="results"></a>

# Results

## Descriptive statistics

<a name="descriptive"></a>

Next we calculate means and standard deviations by condition for the pre and post SSES.

```{r descriptive}
sses_pre %>%
  group_by(condition) %>%
  summarise(mean = mean(total), sd = sd(total))
sses_post %>%
  group_by(condition) %>%
  summarise(mean = mean(total), sd = sd(total))
```

<a name="cronbach"></a>

## Calculate Cronbach’s alpha

See also: [Analysing scales](analyse-scales.html).

We calculate an overall Cronbach's alpha for the pre and post SSES.

```{r cronbach-all}
library(psy)

sses_cronbach <- function(scores) {
  cronbach <- cronbach(select(scores, q1:q20))
  data.frame(cronbach$alpha)
}
sses_cronbach_total <- sses_cronbach(bind_rows(sses_pre, sses_post))
sses_cronbach_total
```

This gives us a value of `0.67`, which is quite a bit lower than the `0.92` reported by the authors who developed the SSES (Heatherton & Polivy, 1991).

We also look at Cronbach's alpha pre and post intervention:

```{r cronbach-pre-post}
sses_pre_cronbach <- sses_pre %>%
  group_by(condition) %>%
  group_modify(~ sses_cronbach(.x))
sses_pre_cronbach

sses_post_cronbach <- sses_post %>%
  group_by(condition) %>%
  group_modify(~ sses_cronbach(.x))
sses_post_cronbach
```

Alpha looks especially low in the `self` and `other` conditions.

<a name="baseline"></a>

## Compare baseline SSES scores between conditions

See also: [Within-subject differences](anova1.html).

We would like to check that there were no major self-esteem differences between conditions before our intervention. This should be the case if we successfully randomised participants to conditions. If there are baseline differences we would need to account for these when comparing them to post-intervention scores. We can use a between-subjects ANOVA to compare the baseline SSES scores in our three conditions.

```{r baseline}
library(BayesFactor, quietly = TRUE)
bf <- anovaBF(formula = total ~ condition, data = as.data.frame(sses_pre))
bf
```

The Bayes factor of `0.68` is greater than the conventional `0.33` which would satisfy us that there were no differences between the conditions. Our descriptive statistics show that the pre-intervention mean and standard deviation are slightly lower in the `other` condition. We'll address this issue in the next section.

```{r baseline-descriptives}
sses_pre %>% group_by(condition) %>% summarise(mean = mean(total), sd = sd(total))
```

<a name="prepost"></a>

# Compare pre and post intervention SSES scores between conditions

See also: [Factorial differences](anova3.html).

The main question we'd like to answer is whether our two imagery interventions 
(visualising a negative mental image of oneself, or someone else) had effects on self esteem which differed from our control condition. We can test this using a [factorial ANOVA](anova3.html#bfact) to compare SSES scores before and after the three interventions. In this design, `condition` is a between-subjects variable, and `time` (pre and post intervention) is a repeated measure.

```{r prepost}
sses_pre_post <- bind_rows(sses_pre, sses_post) %>%
  mutate(time = factor(time))

bf <- anovaBF(formula = total ~ time*condition + subj,
              data = sses_pre_post, whichRandom = 'subj')

```

Our Bayes factors indicate that there is a main effect of `condition`,

```{r condition}
bf[1]
```

no main effect of `time`,

```{r time}
bf[2]
```

and a `condition*time` interaction

```{r condition*time}
bf[4] / bf[3]
```

Given the interaction, we would might want to make [pairwise comparisons](anova1.html#pairs), to compare the imagery conditions against the control condition, and against each other.

Finally, to address the difference in baseline scores, we could run the same ANOVA on the _difference_ between the pre and post scores, or we could include the baseline score as a covariate in an ANCOVA. These techniques are outside the scope of this worksheet.

# References

Heatherton, T. F., & Polivy, J. (1991). Development and validation of a scale for measuring state self-esteem. _Journal of Personality and Social Psychology, 60(6)_, 895.
___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0.