---
title: "The effects of negative mental imagery on self-esteem"
author: "Sophie Homer, Paul Sharpe"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate
## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
options(tibble.width = Inf) # show all columns in output
library(pander)
```

## Before you start

This is an advanced worksheet, which assumes you have completed the [Absolute Beginners' Guide to R](https://ajwills72.github.io/rminr/#beginners) course, the [Research Methods in Practice (Quantitative section)](https://benwhalley.github.io/rmip/overview-quantitative.html) course, and the [Intermediate Guide to R](https://ajwills72.github.io/rminr/#rmip) course.

## Contents

- [Introduction](#intro)
- [Preprocessing](#preproc)
- [Calculating descriptive statistics](#descriptives)
- [Checking SSES reliability](#cronbach)
- [Comparing baseline SSES scores](#baseline)
- [Comparing interventions](#prepost)

<a name="intro"></a>

## Introduction

This worksheet describes a partial analysis pipeline for an experiment taken from a PhD thesis. The experiment explored self-esteem before and after a mental imagery intervention. Self-esteem was measured using the State Self-Esteem Scale (SSES, Osman et al., 2012), a 20-item survey used to measure short-lived (state) changes in self-esteem. Participants completed one of two mental imagery conditions, or a control condition:

 * Condition 1: participants visualised a negative mental image of themself
 * Condition 2: participants visualised a negative mental image of someone else
 * Condition 0 (control): participants did a card sorting task and did not think of any images.

This was a 2 (time) x 3 (condition) mixed design.

<a name="preproc"></a>

## Preprocessing

The initial preprocessing steps for this data are described in the [Data preprocessing for scales worksheet](preproc-scales.html). Complete that worksheet first, **then add the code in this worksheet to the end of `scales.R` and run it**.

```{r preproc, message=FALSE, echo=FALSE}
# from tables.R
rm(list = ls()) # clear the environment
library(tidyverse)

# DASS-21
dass21_raw <- read_csv("going-further/dass21.csv")
dass21_raw <- dass21_raw %>% select(partID, Age:DASS21)
dass21 <- dass21_raw %>% drop_na()
dass21 <- dass21 %>%
  mutate(depression = rowSums(dass21[4 + c(3,5,10,13,16,17,21)]))
dass21_total <- dass21 %>% select(partID, Age, Gender, depression)
dass21 <- mutate(dass21, anxiety = rowSums(dass21[4 + c(2,4,7,9,15,19,20)]))
dass21 <- mutate(dass21, stress = rowSums(dass21[4 + c(1,6,8,11,12,14,18)]))

# SSES
sses <- read_csv('going-further/sses.csv')
sses_pre_raw  <- sses %>% select(1, 5:25)
sses_pre_raw  <- sses_pre_raw %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q"))
sses_pre_raw <- sses_pre_raw %>%
  mutate(subj = factor(partid), condition = factor(condition),
         time = factor('pre')) %>%
  select(subj, condition, time, q1:q20)

sses_pre_raw  <- select(sses, 1, 5:25) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition),
         time = factor('pre')) %>%
  select(subj, condition, time, q1:q20)
sses_post_raw <- select(sses, 1, 25:45) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("post_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition), time = factor('post')) %>%
  select(subj, condition, time, q1:q20)
```

We start with some final bits of preprocessing. **Enter these commands into your script, and run them:**

```{r}
sses_raw <- bind_rows(sses_pre_raw, sses_post_raw)
sses_raw <- sses_raw %>%
  mutate(total = sses_raw %>% select(q1:q20) %>% rowSums())
```

### Explanation of commands

`bind_rows()` simply joins data frames together, in the order the arguments are specificed, making a new data frame. We do this because we want to do at least one analysis on all of the data.

The `mutate()` in the second line uses `rowSums()` to add up the values in the columns `q1:q20` for each row. Because our data has one participant per row, this calculates the SSES score for each participant. You've seen something similar when we calculated subscale scores in the [Data preprocessing for scales](preproc-scales.html#subscales). The SSES contains some reverse-coded items, but these were already reversed in this particular dataset, so we didn't need to do that step.

<a name="descriptives"></a>

## Calculating descriptive statistics

We can get a feel for our data by using the familiar `group_by()` and `summarise()` commands to calculate means and standard deviations by group within time.

**Enter these commands into your script, and run them:**

```{r descriptive}
sses_raw %>%
  group_by(time, condition) %>%
  summarise(mean = mean(total), sd = sd(total))
```

From these results, we can see that, compared to the pre-intervention scores, self-esteem rose slightly in `other` condition, rose even more in the `self` condition and dropped slightly in the `control` condition.

<a name="cronbach"></a>

## Checking SSES reliability

To check the reliability of our SSES measurements, we'll calculate Cronbach's alpha for the pre-intervention SSES data. Cronbach's alpha was introduced in the [Analysing scales worksheet](analyse-scales.html).

**Enter these commands into your script, and run them:**

```{r cronbach-all}
library(psy)
sses_pre <- sses_raw %>% filter(time == 'pre')
sses_pre %>% select(q1:q20) %>% cronbach()
```

### Explanation of commands

Line 1 loads the `psy` package which provides the `cronbach()` function. Line 2 creates a data frame contain only the rows with pre-intervention scores, along with the condition, and SSES columns. We only use the pre-intervention scores because, assuming the interventions were successful, including the post-intervention scores would reduce alpha. The last line calcluates Cronbach's alpha.

### Explanation of output

```{r echo=FALSE}
sses_cronbach_total <- sses_pre %>% select(q1:q20) %>% cronbach()
```

The value of alpha (`r round(sses_cronbach_total$alpha,2)`) could be a cause for concern, as it is below [the 0.7-0.8 convention of acceptable reliability](#analyse-scales.html#cronbach), and well below the alpha of 0.92 reported by the authors of the SSES (Heatherton & Polivy, 1991).

We'll do an additional analysis to see how the scale performed in each group.

**Enter these commands into your script, and run them:**

```{r cronbach-pre-post}
sses_pre %>% filter(condition == 'control') %>%
  select(q1:q20) %>% cronbach()
sses_pre %>% filter(condition == 'self') %>%
  select(q1:q20) %>% cronbach()
sses_pre %>% filter(condition == 'other') %>%
  select(q1:q20) %>% cronbach()
```

### Explanation of commands

Lines 1-2 filter the pre-intervention data to only include the control condition, and then calculates Cronbach's alpha. The remaining lines do the same for the other two conditions.

### Explanation of output

We can now see from these results that alpha was especially low in the `self` and `other` conditions.

<a name="baseline"></a>

## Comparing baseline SSES scores

Putting aside the reliability of the measurements for now, we would like to check that there were no major self-esteem differences between conditions before our intervention. This should be the case if we successfully randomised participants to conditions. If there are baseline differences we would need to account for these when comparing them to post-intervention scores. We can use a between-subjects ANOVA to compare the baseline SSES scores in our three conditions. This is the similar to the ANOVA described in the [Within-subject differences worksheet](anova1.html), but for a between-subjects factor.

**Enter these commands into your script, and run them:**

```{r baseline, warning=FALSE, message=FALSE}
library(BayesFactor, quietly = TRUE)
sses_pre <- sses_raw %>% filter(time == 'pre')
anovaBF(formula = total ~ condition, data = data.frame(sses_pre))
```

### Explanation of commands

In line 1, we load the BayesFactor package. Line 2 assigns just the pre-intervention data to `sses_pre`. Line 3 runs the between-subjects ANOVA. To run [a Bayesian ANOVA using a random factor](anova3.html#bfact), we would need more than one observation for each participant for `condition`. As we only have one observation for each participant, we don't use `+ subj` in the formula, and `whichRandom = 'subj'`.

### Explanation of output

The Bayes factor of .68 is greater than the conventional `0.33` which would satisfy us that there were no differences between the conditions. This could be due to the slightly lower mean in the `other` condition. We'll address this issue at the end of the worksheet.

**Enter these commands into your script, and run them:**

```{r baseline-descriptives}
sses_raw %>% filter(time == 'pre') %>%
  group_by(condition) %>%
  summarise(mean = mean(total), sd = sd(total))
```

<a name="prepost"></a>

## Comparing interventions

The main question we'd like to answer is whether our two imagery interventions 
(visualising a negative mental image of oneself, or someone else) had effects on self esteem which differed from our control condition. We can test this using a [factorial ANOVA](anova3.html#bfact) to compare SSES scores before and after the three interventions. In this design, `condition` is a between-subjects variable, and `time` (pre and post intervention) is a repeated measure.

**Enter these commands into your script, and run them:**

```{r prepost}
bf <- anovaBF(formula = total ~ time*condition + subj,
              data = data.frame(sses_raw), whichRandom = 'subj')
bf
bf[4] / bf[3]
```

### Explanation of commands

The first two lines run a Bayesian factorial ANOVA, with `subj` as a random factor, and store result is stored in `bf`. Line 3 prints the results, which provides us with the Bayes Factors for the main effects of `condition` (`bf[1]`) and `time` (`bf[2]`). Line 4 calculates the Bayes Factor for the interaction.

### Explanation of output

The Bayes Factor for `time` is less than .33, which tells us that there was no overall change in self-esteem after the intervention relative to baseline. The Bayes Factor for `condition` is greater than 3, indicating that there were differences in self-esteem between the three conditions. The final Bayes Factor tells us that it's about 11 times more like that there's an interaction between `condition` and `time` than that there isn't.

# References

Heatherton, T. F., & Polivy, J. (1991). [Development and validation of a scale for measuring state self-esteem](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.476.4746&rep=rep1&type=pdf) _Journal of Personality and Social Psychology, 60(6)_, 895.

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0.