---
title: "The effects of negative mental imagery on self-esteem"
author: "Sophie Homer, Paul Sharpe"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Data required to knit
## https://github.com/ajwills72/rminr-data/tree/master/going-further/sses.csv
##
## I check out rminr-data and make a symbolic link to going-further

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
options(tibble.width = Inf) # show all columns in output
library(pander)
```

## Before you start

This is an advanced worksheet, which assumes you have completed the [Absolute Beginners' Guide to R](https://ajwills72.github.io/rminr/#beginners) course, the [Research Methods in Practice (Quantitative section)](https://benwhalley.github.io/rmip/overview-quantitative.html) course, and the [Intermediate Guide to R](https://ajwills72.github.io/rminr/#rmip) course.

## Contents

- [Introduction](#intro)
- [Preprocessing](#preproc)
- [Calculating descriptive statistics](#descriptives)
- [Checking SSES reliability](#cronbach)
- [Comparing baseline SSES scores](#baseline)
- [Comparing interventions](#prepost)

<a name="intro"></a>

## Introduction

This worksheet describes a partial analysis pipeline for an experiment taken from a PhD thesis. The experiment explored self-esteem before and after a mental imagery intervention. Self-esteem was measured using the State Self-Esteem Scale (SSES, Osman et al., 2012), a 20-item survey used to measure short-lived (state) changes in self-esteem. Participants completed one of two mental imagery conditions, or a control condition:

 * Condition 1: participants visualised a negative mental image of themself
 * Condition 2: participants visualised a negative mental image of someone else
 * Condition 0 (control): participants did a card sorting task and did not think of any images.

This was a 2 (time) x 3 (condition) mixed design.

<a name="preproc"></a>

## Preprocessing

The initial preprocessing steps for this data are described in the [Data preprocessing for scales worksheet](preproc-scales.html). Complete that worksheet first, then add the code in this worksheet to the end of `scales.R`.

```{r preproc, message=FALSE, echo=FALSE}
# from tables.R
rm(list = ls()) # clear the environment
library(tidyverse)

# DASS-21
dass21_raw <- read_csv("going-further/dass21.csv")
dass21_raw <- dass21_raw %>% select(partID, Age:DASS21)
dass21 <- dass21_raw %>% drop_na()
dass21 <- dass21 %>%
  mutate(depression = rowSums(dass21[4 + c(3,5,10,13,16,17,21)]))
dass21_total <- dass21 %>% select(partID, Age, Gender, depression)
dass21 <- mutate(dass21, anxiety = rowSums(dass21[4 + c(2,4,7,9,15,19,20)]))
dass21 <- mutate(dass21, stress = rowSums(dass21[4 + c(1,6,8,11,12,14,18)]))

# SSES
sses <- read_csv('going-further/sses.csv')
sses_pre_raw  <- sses %>% select(1, 5:25)
sses_pre_raw  <- sses_pre_raw %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q"))
sses_pre_raw <- sses_pre_raw %>%
  mutate(subj = factor(partid), condition = factor(condition),
         time = factor('pre')) %>%
  select(subj, condition, time, q1:q20)

sses_pre_raw  <- select(sses, 1, 5:25) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition),
         time = factor('pre')) %>%
  select(subj, condition, time, q1:q20)
sses_post_raw <- select(sses, 1, 25:45) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("post_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition), time = factor('post')) %>%
  select(subj, condition, time, q1:q20)
```

We start with some final bits of preprocessing.

```{r}
sses_raw <- bind_rows(sses_pre_raw, sses_post_raw)
sses_raw <- sses_raw %>%
  mutate(total = sses_raw %>% select(q1:q20) %>% rowSums())
```

### Explanation of commands

`bind_rows()` simply joins data frames together, in the order the arguments are specificed, making a new data frame. We do this because we want to do at least one analysis on all of the data.

The `mutate()` in the second line uses `rowSums()` to add up the values in the columns `q1:q20` for each row. Because our data has one participant per row, this calculates the SSES score for each participant. You've seen something similar when we calculated subscale scores in the [Data preprocessing for scales](preproc-scales.html#subscales). The SSES contains some reverse-coded items, but these were already reversed in this particular dataset, so we didn't need to do that step.

<a name="descriptives"></a>

## Calculating descriptive statistics

We can get a feel for our data by using the familiar `group_by()` and `summarise()` commands to calculate means and standard deviations by group within time.

```{r descriptive}
sses_raw %>%
  group_by(time, condition) %>%
  summarise(mean = mean(total), sd = sd(total))
```

From these results, we can see that, compared to the pre-intervention scores, self-esteem rose slightly in `other` condition, rose even more in the `self` condition and dropped slightly in the `control` condition.

<a name="cronbach"></a>

## Checking SSES reliability

To check the reliability of our SSES measurements, we'll calculate Cronbach's alpha for all of the SSES data. Cronbach's alpha was introduced in the [Analysing scales worksheet](analyse-scales.html).

```{r cronbach-all}
library(psy)
sses_cronbach_total <- sses_raw %>% select(q1:q20) %>% cronbach()
```

This could be a cause for concern, as the value of `r round(sses_cronbach_total$alpha,2)` is below [the 0.7-0.8 convention of acceptable reliability](#analyse-scales.html#cronbach), and well below the alpha of 0.92 reported by the authors of the SSES (Heatherton & Polivy, 1991).

We'll do an additional analysis to see how the scale performed in each group.

```{r cronbach-pre-post}
sses_raw %>%
  group_by(time, condition) %>%
  select(q1:q20) %>%
  group_modify(~ as.data.frame(cronbach(.x)$alpha))
```

### Explanation of commands

The first two steps in this pipeline should be familiar by now. We group the data by `condition` within `time` and select the columns required for calculating alpha. The final step needs some unpacking. The `group_modify()` command does some processing on each group of rows defined by the earlier `group_by()` and returns the result as a data frame, with columns to indicate which group the results relate to. The `~` means turn the following commands into a function which is run for each group. You don't really need to know what a function is to understand the processing that happens for each group. That's defined by `as.data.frame(cronbach(.x)$alpha)`. This runs the `cronbach()` function on the grouped data (`.x`) and extracts the value of `alpha` (recall that `cronbach()` also returns the sample size and the number of items). The alpha value is turned into a data frame so that `group_modify()` can join the results from each group into a single data frame.

### Explanation of output

We can now see from these results that alpha was especially low in the `self` and `other` conditions.

<a name="baseline"></a>

## Comparing baseline SSES scores

Putting aside the reliability of the measurements for now, we would like to check that there were no major self-esteem differences between conditions before our intervention. This should be the case if we successfully randomised participants to conditions. If there are baseline differences we would need to account for these when comparing them to post-intervention scores. We can use a between-subjects ANOVA to compare the baseline SSES scores in our three conditions. This is the similar to the ANOVA described in the [Within-subject differences worksheet](anova1.html), but for a between-subjects factor.

```{r baseline}
library(BayesFactor, quietly = TRUE)
sses_pre <- sses_raw %>% filter(time == 'pre')
bf <- anovaBF(formula = total ~ condition, data = sses_pre)
bf
```

### Explanation of commands

In line 1, we load the BayesFactor package. Line 2 assigns just the pre-intervention data to `sses_pre`. Line 3 runs the between-subjects ANOVA. In contrast with the within-subjects ANOVA you've seen before, we don't use `+ subj` in the formula, and `whichRandom = 'subj'`. For a within-subjects factor, subject is a random factor, because the within-subject factor _levels_ would change if we re-ran the experiment with different people. However, for a fully between-subjects factor like `condition`, the same subject is only ever in a single condition, so it's not possible them to affect more than one level of the factor. This is why we don't include `subj` as a random factor in this analysis.

### Explanation of output

```{r baseline_bf, echo=FALSE}
# calculate here to work around inline knitting issue
baseline_bf <- round(extractBF(bf)$bf,2)
```

The Bayes factor of `r baseline_bf` is greater than the conventional `0.33` which would satisfy us that there were no differences between the conditions. This could be due to the slightly lower mean in the `other` condition. We'll address this issue at the end of the worksheet.

```{r baseline-descriptives}
sses_raw %>% filter(time == 'pre') %>% group_by(condition) %>% summarise(mean = mean(total), sd = sd(total))
```

<a name="prepost"></a>

## Comparing interventions

The main question we'd like to answer is whether our two imagery interventions 
(visualising a negative mental image of oneself, or someone else) had effects on self esteem which differed from our control condition. We can test this using a [factorial ANOVA](anova3.html#bfact) to compare SSES scores before and after the three interventions. In this design, `condition` is a between-subjects variable, and `time` (pre and post intervention) is a repeated measure. We include `subj` as a random factor, as it is random in relation to the repeated measure `time`.

```{r prepost}
bf <- anovaBF(formula = total ~ time*condition + subj,
              data = sses_raw, whichRandom = 'subj')
bf
```

```{r prepost_bf, echo=FALSE}
# calculate here to work around inline knitting issue
condition_bf <- round(extractBF(bf[1])$bf, 2)
time_bf <- round(extractBF(bf[2])$bf, 2)
interaction_bf <- round(extractBF(bf[4] / bf[3])$bf, 2)
```

The Bayes factor for `time` (_BF_ = `r time_bf`) tells us that there was no overall change in self-esteem after the intervention relative to baseline. The Bayes factor for `condition` (_BF_ = `r condition_bf`) tells us that there were overall differences in self-esteem between the three conditions. The Bayes factor of `r interaction_bf`, tells us that it's almost 10 times more like that there's an interaction between `condition` and `time`, than that there isn't.

# References

Heatherton, T. F., & Polivy, J. (1991). [Development and validation of a scale for measuring state self-esteem](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.476.4746&rep=rep1&type=pdf) _Journal of Personality and Social Psychology, 60(6)_, 895.

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0.