---
title: "Exploring data"
author: "Andy Wills"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and ouptut.
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

## Before you start...
Before starting this exercise, you should have had a brief introduction to getting and using RStudio. If not, take a look at the [Introduction to RStudio][1].

[1]: intro-rstudio.html

## Contents

- [How to use these worksheets](#how-to)
- [Loading a package](#package)
- [Loading data](#load)
- [Inspecting data](#inspect)
- [Calculating a mean](#means)
- [Calculating a median](#medians)
- [Dealing with missing data](#missing)
- [Introduction to graphs](#graphs)
- [Customising graphs](#custom-graphs)
- [Lab book writing exercise](#writing)


<a name="how-to"></a>

## How to use these worksheets.

Throughout this worksheet, you'll see the commands you should type into RStudio inside a grey box, followed by the output you should expect to see in one or more white boxes. Any differences in the colour of the text can be ignored.

Each command in this worksheet is followed by one or more  _explanation_ sections - those are there to help you understand how the commands work and how to read the output they produce.

<a name="package"></a>

## Loading a package
First, we need to load a package called *tidyverse*. A package is an extension to R that adds new commands. Nearly everything we'll do in this course uses the tidyverse package, so pretty much every project starts with this instruction. 

**Type (or copy and paste) the command in the grey box into the _Console_ window of RStudio, and press the RETURN key.**


```{r init}
library(tidyverse)
```

When you type the command above and press RETURN, RStudio will print some text to the Console (shown in the white box, above). This text tells you that the _tidyverse_ package has loaded ("attached") some other pacakges (e.g. _dplyr_). It also tells you that the _dplyr_ package changes the way some commands in R work ("conflicts"). That's OK.

If you get an output that includes the word 'error', please see [common errors](common-errors.html).

<a name="load"></a>

## Loading data
Now, we're going to load some data on the income of 10,000 people in the United States of America. I've made up this dataset for teaching purposes, but it's somewhat similar to large open data sets available on the web, such as [US Current Population Survey](https://thedataweb.rm.census.gov/ftp/cps_ftp.html).
```{r load,cache=TRUE}
cpsdata <- read_csv("http://www.willslab.org.uk/cps2.csv")
```

### Explanation of command
There are three parts to the command `cpsdata <- read_csv("http://www.willslab.org.uk/cps2.csv")`:

1. The first part of the command is `cpsdata`. This gives a name to the data we are going to load. We'll use this name to refer to it later, so it's worth using a name that is both short and meaningful. I've called it `cpsdata` because it's somewhat similar to data from the US Current Population Survey, but you can give data pretty much any name you choose (e.g. _fart_). 

2. The bit in the middle, `<-`, is an arrow and is typed by pressing `<` and then `-`, without a space. This arrow means "put the thing on the right of the arrow into the thing on the left of the arrow". 

3. The last part of the command is `read_csv("http://www.willslab.org.uk/cps2.csv")`. It's a way of downloading data from the Internet. The part inside the speech marks, `http://www.willslab.org.uk/cps2.csv`, is a web address, such as you'd use to access any other web page (e.g. <http://www.twitter.com/ajwills72>)

### Explanation of output
R likes to print things in red sometimes -- this does not mean there's a problem. If there's a problem, it will actually say 'error'. The output here tells us that R has loaded the data, which has eight parts (columns, or `cols`). It gives us the name of the columns (`ID, sex, ...`) and tells us what sort of data each column contains: `character` means the data is words  (e.g. 'female'), `integer` means the data is a whole number (e.g. '42').

If you get an error here, please see [common errors](common-errors.html).

<a name="inspect"></a>

## Inspecting data
Next, we'll take a peek at these data. You can do this by clicking on the data in the *Environment* tab of RStudio, see [Introduction to RStudio](intro-rstudio.html).

We can now see the data set (also known as a *data frame*). We can see that this data frame has 8 columns and 10000 rows. Each row is one person, and each column provides some information about them. Below is a description of each of the columns. Where you see `NA` this means this piece of data is missing for this person -- quite common in some real datasets.

Here's what each of the columns in the data set contains:

| Column | Description                             | Values             |
| ------ | --------------------------------------- | ------------------ |
| ID     | Unique anonymous participant number     | 1-10,000           | 
| sex    | Biological sex of participant           | male, female       |
| native | Participant born in the US?             | foreign, native           |
| blind  | Participant blind?                      | yes, no            |
| hours  | Number of hours worked per week         | a number           |
| job    | Type of job held by participant:        | charity, nopay, private, public |
| income | Annual income in dollars                | a number           |
| education | Highest qualification obtained       | grade-school, high-school, bachelor, master, doctor |

<a name="means"></a>

## Calculating a mean
Now we have these data, one question we can ask is "what is the average income of people in the U.S.?" (or, at least, in this sample). In this first example, we're going to calculate the _mean_ income. 

I'm sure you learned about means in school but, as a reminder, you calculate a mean by adding up all the incomes and dividing by the number of incomes. Our sample has 10,000 participants, so this would be a long and tedious calculation -- and we'd probably make an error. 

It would also be a little bit tedious and error prone in a spreadsheet application (e.g. Excel, Libreoffice Calc). There are some very famous cases of these kinds of "Excel errors" in research, e.g. [genetics](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-016-1044-7), [economics](https://arstechnica.com/tech-policy/2013/04/microsoft-excel-the-ruiner-of-global-economies/).

In R, we can calculate the mean instantly, and it's harder to make the sorts of errors that are common in Excel-based analysis.

To calculate mean income in  R, we type the following command:

```{r mean}
cpsdata %>% summarise(mean(income))
```

Your output will tell you the mean income in this sample -- it's the last number on the bottom right, and it's approximately $87,000.

If you're happy with the output you've got, move on to the next section. If you would like a more detailed explanation of this output, see [more on tibbles](more-on-tibbles.html). 

If you get an error here, please see [common errors](common-errors.html).

**RECORD YOUR ANSWER** - Type the exact mean income into the Answer Section of Psyc\:EL.

### Explanation of command
This command has three components:

1. The bit on the left, `cpsdata`, is our data frame, which we loaded and named earlier. 

2. The bit in the middle, `%>%`, is called a *pipe*. Its job is to send data from one part of your command to another. It is typed by pressing `%` then `>` then `%`, without spaces. So `cpsdata %>%` sends our data frame to the next part of our command.

3. The bit on the right, `summarise(mean(income))` is itself made up of parts. The command `summarise` does as the name might suggest, it summarises a set of data (`cpsdata` in this case) into a single number,  e.g. a mean. The `mean` command indicates that the type of summary we want is a mean (there are also a number of other types of summary, as we'll see later). Finally, `income` is the name of the column of `cpsdata` we want to take the mean of -- in this case, the income of each individual.

<a name="medians"></a>

## Calculating a median
Now we're going to calculate the median income of the people in this sample. As you learned in school, you calculate a median by putting all the numbers into rank order and then picking the number in the middle. As with the calculation of mean outcome, R allows us to calculate the median quickly and without error.

This time, I haven't given you the command you need to type -- your task is to work out what you need to type. Re-read the explanation above for clues if you need them. The way to indicate that the summary you want is a median is to use the command `median`. 

**TIP**: You can get back to commands you've typed previously by pressing the up arrow on your keyboard while you're in the Console. This way, you can get the command for calculating a mean, and edit it to calculate a median.

If you've entered the correct command, you'll find that the  median income is $56,952.
```{r median, echo=FALSE}
cpsdata %>% summarise(median(income))
```

**RECORD YOUR ANSWER** - Type the exact command you used to calculate median income into the Answer Section of Psyc\:EL.

<a name="missing"></a>

## Dealing with missing data

To calculate the mean number of hours worked per week, we have to deal with the fact that there is some missing data - we don't know for all 10,000 people how many hours they work in a week, because they didn't all tell us. To get a mean of those who did tell us, we tell R to ignore the missing data, like this:

```{r mean-hours-narm}
cpsdata %>% summarise(mean(hours, na.rm = TRUE))
```

### Explanation
`rm` is short for 'remove', but 'ignore' would be a more accurate description, as this command doesn't delete the `NA` entries in `cpsdata`, it just ignores them. So `na.rm = TRUE` means "ignore the missing data". 

If you get an error here, please see [common errors](common-errors.html).

<a name="graphs"></a>

## Introduction to graphs
In the last two exercises, we found that median US income was much lower than mean US income. To help us explore why that might be, we're going to look at the *distribution* of incomes. Again, we're going to use a concept you learned in school -- we're going to produce a histogram.

A histogram is a graph that shows us how many people have an income that is within a number of equal, consecutive, ranges (also called *bins*). In this example, we're going to count how many people earn $0-19,999, how many earn $20,000-39,999, and so on, until we reach the highest income in the sample. So, our *bin width* will be $20,000. We're then going to represent these counts by the height of a series of bars.

In R, we can do this with a single command:
```{r histo}
cpsdata %>% ggplot(aes(income)) + geom_histogram(binwidth=20000)
```

### Explanation of command

1. The first part, `cpsdata %>%` works the same way as in the previous examples. It takes the data from `cpsdata` and *pipes* it to our graphing command `ggplot`.

2. The command `ggplot` needs to know which of the columns you want to show on your graph, and we use the command `aes` (short for "aesthetics") to specify this. In this case, we want to plot the incomes, so the command is ``ggplot(aes(income))``

3. We also have to tell R *how* we want to graph the income data. The command ``geom_histogram`` says that we want a histogram (`geom_` in this context just means *graph*). We also need to specify the *bin width* for our histogram, using `binwidth`. 

<a name="custom-graphs"></a>

## Customising graphs

There are a lot of ways to modify the standard graphs produced by R, to make them look the way you want. Here, we're going to try just a few of them. 

### Changing the theme
We can change the overall look of a graph by changing the *theme*. In this example, we use a **lighter background** by adding the command `+ theme_light()`:
```{r themes,fig.keep='none'}
cpsdata %>% ggplot(aes(income)) + geom_histogram(binwidth=20000) + theme_light()
```

There are quite a lot of different themes. Try replacing `light` above with one of the following: `bw, classic, gray, linedraw, light, minimal, void, dark` 

### Changing the colour
We can also change the way particular parts of the graph look -- for example, changing the colour of the bars on the histogram, using the command `fill`. Here's one particularly nasty-looking example:

```{r colours}
cpsdata %>% ggplot(aes(income)) + geom_histogram(fill = 'yellow', binwidth=20000) + theme_dark()
```

Try replacing `yellow` with some other colour name. R knows quite a lot of colour names.

### Changing the labels
We can also change the labels that appear on the x-axis and on the y-axis. Here's an example with some not-very-helpful labels:
```{r labels}
cpsdata %>% ggplot(aes(income)) + geom_histogram(binwidth=20000) + 
  xlab('insert a x-axis label here') + ylab('insert a y-axis label here')
```

### Lab book exercise - Upload a graph
Generate a histogram of these data with a bin width of $50,000. The histogram bars should be blue, and you should use `theme_bw`. Give your x-axis and y-axis meaningful labels. If you get it right, your graph should look something like this (without the words "EXAMPLE PLOT", of course):
```{r hist-assess, echo=FALSE, fig.keep='all'}
cpsdata %>% ggplot(aes(income)) + geom_histogram(fill = 'blue', binwidth=50000) + theme_bw() +
  xlab('Income in US dollars') + ylab('Number of people') + 
  annotate("text", x = Inf, y = -Inf, label = "EXAMPLE PLOT", col = "red", hjust = 5.1, vjust = -1.1)
```

Now, export your histogram, using the _Export_ icon on RStudio's _Plots_ window, and selecting "Save as image...". Give it a meaninful file name (e.g. "ExploringIncomes") and click 'Save'. 

**RECORD YOUR ANSWER** - _Psyc\:EL_ and _RStudio Online_ are entirely separate web pages that don't talk to each other, so you'll need to download your graph from RStudio Online and then upload it to Psyc\:EL. Here's how:

### Downloading files from RStudio Online

1. Click on the 'Files' tab in RStudio Online, and click the tick box to select your image (it will have whatever filename you just gave it.). 

2. Click on "More" (next to the little blue gear wheel), and then "Export...". 

3. Click "Download". The file will now be in the Downloads folder of your computer. 

### Uploading files to PsycEL

4. Open your computer's Downloads folder (e.g. on a Windows 10 machine, click on the little yellow folder at the bottom of the screen, and then click on "Downloads"). 

5. Find your file in that Downloads folder, and drag it to the appropriate box on your Psyc\:EL Answer Sheet. 

6. Click 'Upload image' on your Psyc\:EL answer sheet. 

### Extension exercise

Take a look at the [R Graph Gallery](https://www.r-graph-gallery.com/) for lots of other examples of how to make graphs in R. Make a pretty graph of some aspect of this data set that interests you.

<a name="writing"></a>

## PsycEL writing exercise

In class, we discussed how the histogram helped us to understand why the mean and median incomes were so different, and whether the mean or the median gave a better account of average US income. Here are the [slides](mean-median.html), if you'd like to look at them again.

On the Psyc\:EL Answer Section, you'll find the question "Does the mean or the median give a better indication of average salary in this case?". **In the text box provided, give a short answer to this question**. You need not write more than a few sentences, but you should explain your reasoning. 

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 
