---
title: "Face recognition"
author: "Chris Longmore, Andy Wills"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and ouptut.
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

## Before you start...
Before starting this exercise, you should have completed **all** the [Absolute Beginners'](index.html) workshop exercises. If not, take a look at those exercises before continuing. Each section below also indicates which of the earlier worksheets are relevant.

## Getting the data into R
**Relevant worksheet:** [Using RStudio projects](using-projects.html)

In order to complete this worksheet, you'll need to have downloaded your CSV file from the PsycEL exercise. **See the instructions on PsycEL for how to do this.**

Once you have downloaded your CSV file, [set up a new project](using-projects.html) on RStudio Server for this analysis, and [upload your CSV](entering-data-by-hand.html#upload) to your project.

## Exploring your data

### Load
**Relevant worksheet:** [Exploring data](exploring-incomes.html) 

Load the _tidyverse_ package, and load your data. 
**Note**: Everyone's CSV file has a different name. For example, yours might be called `10435678 trials.csv`. In the example below, you'll need to replace `facerec.csv` with the name of your personal CSV file.

```{r init-load, message=FALSE} 
library(tidyverse)
recog <- read_csv("facerec.csv")
```

### Inspect
Look at the data by clicking on it in the _Environment_ tab in RStudio. Each row is one trial of the final test. Here's what each of the columns in the data set contain:

| Column    | Description                             | Values             |
| --------- | --------------------------------------- | ------------------ |
| StudentID | Your Student Reference Number           |        |
| TrialNum  | Trial number                            | 0 - 47 |
| Stimulus  | Picture shown on this test trial. Each picture has a unique code that identifies it. The number identifies the person, and the letter identifies which photograph it is of that person.  | e.g. 23B, 4A    |
| Condition | Type of picture shown on this test trial | NS =  person not shown during the study phase; SO = person shown during study, same photo; SD = person shown during study, different photo |  
| Correct   | Did you get this trial correct?        | 1 = Yes, 0 = No  |


### How accurate were you?

**Relevant worksheet:** [Exploring data](exploring-incomes.html). 

Calculate your overall accuracy in this task -- the proportion of trials you got right.The `Correct` column codes a correct response as `1` and an incorrect response as `0`. This is helpful, because the mean of this column gives the proportion of trials you got right. 

```{r accuracy}
recog %>% summarise(mean(Correct))
```

Your output will look something like the above, but the actual number won't be the same. Enter the answer calculated from **your own data** in to your lab book. 

### Old faces only
**Relevant worksheet:** [Group differences](group-differences.html). 

These kinds of experiments typically focus on the faces that were presented during training (the "old" faces), so we're going to filter out the trials on which you were shown a new person -- the `NS` trials. 

We have to tell the `filter` which trials we want to _keep_, so the filter we need is `Condition != 'NS'` where `!=` means 'not equal to'. We put this filtered data into a new data frame, which we've called `recog.old` (because it only contains "old" faces, i.e. the ones seen during training).

```{r filter}
recog.old <- recog %>% filter(Condition != "NS")
```

### Does having the same photograph help?
**Relevant worksheet:** [Group differences](group-differences.html). 

Were you more accurate when the photograph was the same as the one you studied, than when it was a different photograph of the same person? To find out, we `group_by` the type of trial (`Condition`) and then calculate the mean score for each group. 

```{r compare-means}
grpmeans <- recog.old %>% group_by(Condition) %>% summarise(Correct = mean(Correct))
```

We've put our answers into a new data frame, `grpmeans`, so go to the _Enviornment_ window of RStudio and click on `grpmeans' to see your answers.

**Enter your answers  in to your lab book. **

Notice that we used `summarise(Correct = mean(Correct))`, rather than just `summarise(mean(Correct))`. This causes that column of our data frame to be called `Correct`, which is less of a mouthful than `mean(Correct)`. The graphing command we'll use next only works if the columns have very simple names (a single word, no special characters like brackets, etc.)

### Making a bar chart

**Relevant worksheet:** [Exploring data](exploring-incomes.html). 

Next, let's make those two numbers into a bar chart. We didn't cover bar charts in the _Absolute Beginners' Guide_, so here's how to do one:

```{r bar-chart}
meanplot <- grpmeans %>% ggplot(aes(x = Condition, y = Correct)) + geom_col()
meanplot
```

Your output will look a bit like the above, but the heights of the bars will probably be different. 

#### Explanation of command

`meanplot <- ` - This gives the bar chart a name (`meanplot`), so we can modify it later without having to retype everything.

`ggplot` is the command in R used to draw graphs

`aes` tells ggplot which columns of your data frame to use in the plot. In this case, we want 'Condition' on the x axis, and 'Correct' (i.e. proportion correct) on the y axis. 

`+ geom_col()` says that the type of graph we want is a bar chart (`col` is short for columns, another word for bars). 

`meanplot` - This tells R to display the graph that we have named `meanplot`.

### Labelling a chart

**Relevant worksheet:** [Exploring data](exploring-incomes.html). 

Next, give your bar chart a more appropriate label for the y-axis. "Correct" isn't very informative, "Mean proprotion correct" would make more sense. So, we'll use the `ylab` command to add this y-axis label.

```{r bar-chart-label}
meanplot <- meanplot + ylab("A more meaningful label")
meanplot
```

If you get it right, it'll look something like this -- but of course with your meaningful label, rather than the one you see here.

### APA style

**Relevant worksheet:** [Exploring data](exploring-incomes.html). 

The graph you now have is readable, but it doesn't look much like the graphs you seen in journal articles. In the [Exploring data](exploring-incomes.html) worksheet, we saw that we could change the way graphs looked using _themes_; for example, `theme_bw()`. In this exercise, we're going to use a theme that makes graphs look like they do in journal articles. It's called `theme_APA` because it's based on how the Amercian Psychological Association says graphs should look.

The APA theme is not part of any R package at the moment, but you can download it from my website like this:

```{r theme-apa-load}
source("http://www.willslab.org.uk/rminr/theme-apa.R")
```

Now all you need to do is add this theme to your graph, using `+ theme_APA()`. If you get it right, your graph will look something like this (without the words "example plot", of course, and with your meaningful y-axis label):

```{r theme-apa-apply, echo=FALSE}
meanplot <- meanplot + theme_APA() + annotate("text", x = Inf, y = -Inf, label = "EXAMPLE PLOT", col = "red", hjust = 5.1, vjust = -1.1)
meanplot
```

**Use RStudio to export your graph as an Image, and upload it to your lab book.**

### Recording our answer

1. Export your graph, using the _Export_ icon on RStudio's _Plots_ window, and selecting "Save as image...". Give it a meaninful file name (e.g. "face-bar") and click 'Save'. 

2. Download your graph from RStudio server - see [these instructions](using-projects.html#download) for a reminder of how to do this.

3. **Upload your graph** to PsycEL (see the PsychEL activity for instructions of how to do this).


___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

_version 0.2_