---
title: "Animal welfare"
author: "Phil Gee, Andy Wills"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and ouptut.
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

## Before you start...
Before starting this exercise, you should have completed **all** the relevant [Absolute Beginners', Parts 1 & 2](index.html) worksheets. Each section below indicates which of the earlier worksheets are relevant.

## Getting the data into R

**Relevant worksheet:** [Using RStudio projects](using-projects.html)

You and your partner must first complete the behaviour coding exercise. You'll then get a CSV file that contains both your ratings. [Set up an RStudio project](using-projects.html) for this analysis, and [upload the CSV file](entering-data-by-hand.html#upload) to it. 

## Exploring your data

### Load
**Relevant worksheet:** [Exploring data](exploring-incomes.html) 

Load the _tidyverse_ package, and load your data.

```{r init-load, message=FALSE} 
library(tidyverse)
animals <- read_csv("animals.csv")
```

**Note**: Everyone's CSV file will have a different name. In the example above, you'll need to replace `animals.csv` with the name of your personal CSV file.

### Inspect
Look at the data by clicking on it in the _Environment_ tab in RStudio. Each row is one time point in the video you coded. Here's what each of the columns in the data set contain:

| Column    | Description                             | Values             |
| --------- | --------------------------------------- | ------------------ |
| time      | Time point.           | 1 - 10        |
| period    | How long before feeding time? (in minutes). | 10 or 180 | 
| behav.r1 | Rater 1's coding of the animal's behaviour at each time point. | In the example file, you'll find: "pacing", "sleeping", "standing", "lying", "running". Your codes may be different. |
| behav.r2 | Rater 2's coding of the animal's behaviour at each time point. | as above. |
| loc.r1    | Rater 1's coding of the animal's location at each time point. |  In the example file, you'll find: "zone_1", "zone_2", "zone_3", "zone_4". Your codes may be different. |
| loc.r2    | Rater 2's coding of the animal's location at each time point. |  as above. |


## Percentage agreement

**Relevant worksheet:** [Inter-rater reliability](irr.html) 

To what extent did you and your workshop partner agree on how each behaviour should be coded? As we covered in the _inter-rater reliability_ worksheet, to look at this, we first have to `select` the relevant columns of the data frame. For example, to look at inter-rater reliability for the _behaviour_ category, we select:

```{r select, message=FALSE} 
behav <- animals %>% select(behav.r1, behav.r2)
```

We can now use the `agree` command to work out percentage agreement:

```{r pc, message=FALSE} 
library(irr)
agree(behav)
```
**NOTE**: If you get an error here, type `install.packages("irr")`, wait for the pacakge to finish installing, and try again. 

The key result here is `%-agree`, which is your percentage agreement. The term `Subjects` here is a bit misleading, it doesn't mean the number of animals you observed (this data file contains your ratings of one animal), it means the number of time points you recorded an observation for. 

**Enter the percentage agreement for your behaviour and location codings into PsycEL**.

## Cohen's kappa

**Relevant worksheet:** [Inter-rater reliability](irr.html) 

One problem with the _percentage agreement_ measure is that people will sometimes agree purely by chance. Jacob Cohen thought it would be much neater if we could have a measure of agreement where zero always meant the level of agreement expected by chance, and 1 always meant perfect agreement. To calculate his mesaure, Cohen's kappa, in R we use the command `kappa2`: 

```{r ckauto}
kappa2(behav)
```

**Enter the Cohen's kappa values for your behaviour and location codings into PsycEL.**

There are some words that psychologists sometimes use to describe the level of agreement between raters, based on the value of kappa they get. These descriptions are listed in the [inter-rater reliability](irr.html) worksheet, in the section "Describing Cohen's kappa".

**On PsycEL, select the correct term to describe the kappa values for your behaviour and location codings.** 

If either of those descriptions are 'moderate' or lower, reflect on why that might be. For example, is there a problem with the definitions of the behavioural categories you used? What else might have caused the lack of agreement? 

**Write a few sentences into PsycEL summarising your reflections.** 


## Behaviour and feeding time

**Relevant worksheet:** [Relationships](relationships.html) 

Does the lion behave differently when it’s close to feeding time? To look at
this, we need to calculate the frequency of each behaviour at our two time
periods (10 minutes before feeding, and 180 minutes before feeding). You can
use the `table` command we learned in the _Relationships_ worksheet to do this, but you’re going to have to choose which behaviour, and which of your two raters,
to look at.  That's because it's likely you will have had at least a few
disagreements.  But if both of you were looking at the same behaviour, how can
we decide who was 'right'?  There are a few possible solutions, but for now we
will take the simplest: flip a coin to decide which of your raters' data you
will use.

If you choose "behaviour" and rater 2, the commands would be:

```{r cont}
cont <- table(animals$period, animals$behav.r2)
cont
```

### Explanation of command

What you have just done here, as we covered in the [relationships](relationships.htm#cont) worksheet, is to convert your _data frame_, called  `animals`, into a _contingency table_, called `cont`. This contingency table shows how often each behaviour occurs at each time period. Recall that `table(rows, columns)` is the command used in R for producing contingency tables. We replace the word `rows` with the name of the variable we want to appear on the rows of the table, and we replace the word `columns` with the name of the variable we wnat to appear in the columns of the table.

### Bar chart

**Relevant worksheet:** [Face recognition](face-recog.html) 

To visualise the relationship between behaviour and feeding time, we're going to use a bar chart. We covered bar charts in the _Face recognition_ worksheet; here we're going to extend that example to create a bar chart that shows our two different time periods on the same axes. 

```{r bars, fig.show='hide'}
df <- data.frame(cont)
colnames(df) <- c("Period", "Behaviour", "Frequency")
df %>% ggplot(aes(x = Behaviour, y = Frequency, fill = Period)) + geom_col(position="dodge")
```

### Explanation of command

This graph command goes a bit beyond what we've covered in previous worksheets, so here's an explanation of how the new bits work:

`df <- data.frame(cont)` - A _data frame_ is the standard way R stores data (e.g. `animals` is a data frame). The `ggplot` commands expects to get a data frame, and gets upset if it gets something else, like a contingency table. So, the first thing we do is make a data frame version of `cont` (our contingency table), and give it a name (`df` in this case). 

If you click on `df` in the _Environment_ tab of RStudio, you'll see that the rows of the contingency table have been called "Var1" and the columns have been called "Var2". These are not very meaningful labels, so we use the `colnames` command (short for "column names") to give them more meaningful names. This will make our graph clearer. We do this using the command: `colnames(df) <- c("Period", "Behaviour", "Freq")`.

`ggplot(aes(x = Behaviour, y = Freq, fill = Period))` - As in previous bar graphs you've made, you need to tell `ggplot` which data is on the x axis, and which is on the y axis. The new bit here is that we also tell `ggplot` to produce two different colours of bars, with the colour depending on `Period`. 

`geom_col(position="dodge")` - As before, `geom_col` is the command for a "column" plot (aka. a bar chart). The new part here is `position="dodge"`; this tells `ggplot` that you want the two different colours of bars to be placed side-by-side, rather than directly on top of each other (i.e. you want them to "dodge" each other).


### Describing your results

```{r bars-eg, echo=FALSE}
df <- data.frame(cont)
colnames(df) <- c("Period", "Behaviour", "Frequency")
df %>% ggplot(aes(x = Behaviour, y = Frequency, fill = Period)) + geom_col(position="dodge") +
annotate("text", x = Inf, y = -Inf, label = "EXAMPLE PLOT", col = "red", hjust = 3.1, vjust = -1.1)

```

In the above example graph, we can see that the animal was pacing and standing
as often in the two time periods, but was lying slightly more 10 minutes before
feeding time than 180 minutes before feeding time, and sleeping slightly more 180
minutes before feeding time than 10 minutes before. What do your data show?
Did proximity to feeding time have an effect on behaviour?  If so, which
behaviours were most affected?

**Enter a few sentences into PsycEL describing what your data show.**


### Uploading your graph

1. Export your graph, using the _Export_ icon on RStudio's _Plots_ window, and selecting "Save as image...". Give it a meaninful file name (e.g. "feed-time") and click 'Save'. 

2. Download your graph from RStudio server - see [these instructions](using-projects.html#download) for a reminder of how to do this.

3. **Upload your graph to PsycEL** (see the PsychEL activity for instructions of how to do this).



___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0.. 




