---
title: "Calculating your module mark"
author: "Peter Jones, Andy Wills"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and ouptut.
knitr::opts_chunk$set(echo = TRUE, comment=NA)
```

You can use R to quickly calculate your final mark for a module from your component marks, assuming you know the weighting of each component (usually provided on the module description). Here's how:

1. Start a new project in RStudio.

2. Enter your percentage marks for each component for the module:

```{r enter-marks}
portfolio <- 65
poster <- 22
mcq <- 87
data <- 35
```

Of course, replace the marks above with your own marks :-)

3. Combine your marks into one variable using the `c` command:

```{r components <- c(portfolio, poster, mcq, data)bine}
components <- c(portfolio, poster, mcq, data)
```

4. Enter the weighting for each component (same order as above):

```{r weights}
weights <- c(.35, .3, .3, .05)
```

5. Calculate the weighted sum of your marks, using the `sum` command, and then round to the nearest whole percentage, using the `round` command:

```{r mark}
round(sum(components * weights))
```

The number you get from the above command is your overall module mark. 

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

