---
title: "More on data"
author: "Paul Sharpe"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA)
library(pander)
```

# Loading data from Excel

Sometimes the data at the start of your preprocessing pipeline will be an Excel file, rather than a CSV file. As a proprietary format, this is not ideal. However, people often use Excel to keep related data in a single file, by using "subsheets". Whilst you could save each subsheet to its own CSV file, `R` can also read the data directly from Excel.

To demonstrate this, we'll read the data from an experiment where the results of four different tests were recorded on subsheets of an Excel spreadsheet. There was also a subsheet containing demographic data.

We'll read data from all of the subsheets into a single data frame.

```{r excel, message=FALSE}
rm(list = ls()) # clear the environment
library(tidyverse)

# read data from Excel
library(readxl)
path <- 'case-studies/allegra-cattani/picture-naming.xlsx'
data <- path %>%
  excel_sheets() %>%
  set_names() %>%
  map_df(~ read_excel(path = path, sheet = .x, range = "A1:V20"),
         .id = "sheet")
```

**Explanation of commands:**

1. We clear the workspace and load `tidyverse`.
1. `library(readxl)` loads a package containing functions for reading Excel spreadsheets.
1. `path <- 'going-further/picture-naming.xlsx'` defines the location of our spreadsheet.
1. `path %>% excel_sheets()` creates a vector (R calls this type of list a vector) containing the names of the subsheets.
1. Items in a list can be given names. Without an argument, `set_names()` sets the name the same as the value in the list, thereby naming each item after its subsheet. It will become clear why we do this in the next step.
1. `map_df()` processes each item in the list, by assigning it to `.x` and then running the function `read_excel(path = path, sheet = .x, range = "A1:V20")`. For each subsheet of the spreadsheet file (`path`), this reads data from the range `A1:V20` on the subsheet (the cells containing the data we need). The argument `.id = "sheet"` creates a column named `sheet` with the name associated with the current list item `.x`. This is the name that we set in the previous step. The results are returned as a data frame.
1. Because `map_df()` adds the results of each of function call to the end of the previously created data frame, `data` is assigned a single data frame containing the data from all subsheets.

Here's a selection of rows from `data`:

```{r echo=FALSE}
data %>% slice(1:2, 20:21, 39:40, 58:59, 77:78) %>% pander(split.table = Inf)
```


___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 
