---
title: "More on Bayes Factors"
author: "Andy Wills"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and ouptut.
knitr::opts_chunk$set(echo = TRUE, comment=NA)

library(tidyverse)
library(BayesFactor)
cpsdata <- read_csv(url("http://www.willslab.org.uk/cps2.csv"))
cpslow <- cpsdata %>% filter(income < 150000)
```

In the [Evidence](evidence.html) worksheet, we did the following Bayesian t-test:

```{r bfttest}
ttestBF(formula = income ~ sex, data = data.frame(cpsdata))
```

Here's a more detailed explanation of the output of that test -- we'll go through each bit in turn:

`Bayes factor analysis` - You're doing a Bayes Factor analysis.

`[1] Alt., r=0.707` - In order to calculate a Bayes Factor, R has to make some assumptions about how big the difference is likely to be. The `ttestBF` command makes some broad assumptions that cover the range of effect sizes typically seen in psychology. 

More specifically, `ttestBF` assumes a Cauchy distribution of effect sizes centered on zero, with a scale of 0.707 -- this is where `r=0.707` comes from. That description probably didn't make much sense useless you have a very strong maths background, so here's the same idea, shown as a density plot. It's basically an assumption that small effect sizes are quite likely, and large effect sizes are quite unlikely.

```{r cauchy, echo=FALSE}
x<-seq(-10,10,by=0.1)
plot(x, dcauchy(x, scale = .707),type="l", xlab = "Effect size", ylab = "")
```

`18.25128` - The Bayes Factor (i.e. the main result of this analysis)

`±0%` - Basically a confidence interval on the Bayes Factor. It's 0% here because we have so much data, but with smaller samples we might see something like `20 ±5%`, which would means the Bayes Factor is about 20, give or take 5%. So, it's between 19 and 21.

`Against denominator:` - This tells you that the null hypothesis is the denominator of the fraction that is used to calculate the Bayes Factor. So, in other words, you're getting BF~10~ rather than BF~01~ -- see the [Evidence](evidence.html) worksheet.

`Null, mu1-mu2 = 0` - This tells you that the null hypothesis is that the difference between the two group means (called `mu1` and `mu2` here) is zero. 

`Bayes factor type: BFindepSample` - You're doing a Bayes Factor analysis; `indepSample` is short for `independent samples`, which is another way of saying "between-subjects test". `JZS` is short for Jeffreys, Zellner, and Siow - the surnames of three people who are credited with coming up with this particular way of working out Bayes Factors.

___


#### Licence
This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

___

_version: 0.2_
