---
title: "More on Cohen's Kappa"
author: "Andy Wills"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and ouptut.
knitr::opts_chunk$set(echo = TRUE, comment=NA)
library(tidyverse)
library(irr)
friends <- read_csv("irr_nan.csv")
```
In this brief extension worksheet, we look at why kappa is sometimes much lower than percentage agreement, and also why the `kappa2` command sometimes prints `NaN` for _Z_ and _p_. 

To illustrate these things, here are some example ratings, and the output they produce:

```{r init-load, message=FALSE, echo=FALSE}
friends
ratings <- friends %>% select(rater1, rater2)
agree(ratings)
kappa2(ratings)
```


## Why is kappa zero in this case?

It might seem odd to have a kappa of zero here, because the percentage agreement is quite high (80%). Recall that Cohen's kappa is calculated as:

_(P - C) / (100 - C)_

where _P_ is the percentage agreement between the two raters, and _C_ is the percentage agreement we'd expect by chance. So, for kappa to be zero, the percentage agreement by chance must also be 80%. 

Agreement by chance is so high here because Rater 1 is using the same response all the time, and Rater 2 is using that same response 80% of the time. If one person always makes the same rating, and the other makes that rating on a random 80% of occasions, they'll agree 80% of the time. For example, if I call everything I see a cat, and you call everything you see a cat unless you roll a five on your five-sided dice, we'll agree 80% of the time. This does not mean either of us knows what a cat is.

## Why are Z and p equal to `NaN`

In this case, `NaN` doesn't mean grandmother, it means 'not a number'. What's happened here is that there is so little variation in the ratings used (they are nearly all '3') that R cannot calculate the Z score or the p value --- the calculations that are used to do this break down in these extreme cases.


----

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 



