---
title: "More on regression"
author: "Paul Sharpe and Andy Wills"
output:
  html_document:
    highlight: pygment
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate
## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)
## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
library(pander)
# Line numbering guide
# https://blog.atusy.net/submodules/rmd-line-num/index.html
```


## Introduction

You already know about linear regression on a single variable. Multiple regression with two variables. Straightforward to add more variables.

In this worksheet we'll use data from an undergraduate student dissertation which explored relationships between Polish (n=93) and British (n=104) people's emotional intelligence, religious orientation and spirituality.

197 participants (93 polish, 104 british) were analysed (exclusions: 10 missing data, 2 no consent) Polish/English.


<a name="load"></a>

## Loading data

Open the `rminr-data` project we used [previously](preproc.html#load).

Ensure you have the latest files by asking git to "`pull`" the repository. Select the `Git` tab, which is located in the row of tabs which includes the `Environment` tab. Click the `Pull` button with a downward pointing arrow. A window will open showing the files which have been pulled from the repository. Close the `Git pull` window. The `going-further` folder should contain the file `religion.csv`.

Next, create a new, empty R script and save it in the `rminr-data` folder as `more-regression.R`. Put all the commands from this worksheet into this file, and run them from there. Save your script regularly.

We start by loading the data.

```{r load, message=FALSE}
rm(list = ls()) # clear the environment
library(tidyverse)

data <- read_csv('going-further/religion-preproc.csv')
```

### Demographics

```{r}
data <- data %>%
  mutate(sex = factor(sex), education = factor(education))
data$education <- fct_relevel(data$education, 'no formal quals', 'GCSE or equiv', 'A level or equiv', 'Technical, HNC, FD or equiv', 'degree')
```

| Column | Description                             | Values             |
| ------ | --------------------------------------- | ------------------ |
| subj   | Unique anonymous participant number     | 1-197              | 
| age    | Age of participant                      | 18-74              |
| sex    | Biological sex of participant           | male, female       |
| education | Highest education level of participant  | no formal quals, GCSE or equiv, A level or equiv, degree, Technical, HNC, FD or equiv |
| religious  | Did the participant identify with a recognised religion? | yes, no  |

We now turn to the three scales that the participants completed. In practice, you would need to calculate the scores for each scale from the raw data. This is explained in the [Preprocessing scales worksheet](preproc-scales.html). In the the data we loaded, these scores have already been calculated.

### Religions Orientation
 
Religious orientation (RO) was was measured with the amended Religious Orientation Scale (ROS; Maltby & Lewis, 1996). According to the ROS, people who are intrinsically religious treat religion as a spiritual end in and of itself. Those who are extrinsically religious practise religion for self-serving reasons, such as social status. Therefore, the ROS has two subscales. 

| Column | Description                    | Values |
| ------ | ------------------------------ | ------ |
| ro_i   | Intrinsic relgious orientation | 1-3    | 
| ro_e   | Extrinsic relgious orientation | 1-3    |


### Spirituality

Spirituality was measured with the Spiritual Connection Questionnaire (SCQ-48; Wheeler and Hyland, 2008). The SCQ-48 has five factors: connection with nature, connection with places, connection with the universe, connection with other people, and .

| Column | Description                                                            | Values |
| ------ | ---------------------------------------------------------------------- | ------ |
| happiness | Extent to which spirituality brings the participant happiness | 1-7| 
| places   | Extent to which the participant feels spiritually connected to places | 1-7 |
| others | Extent to which the participant feels spiritually connected to others | 1-7  |
| nature | Extent to which the participant feels spiritually connected to nature | 1-7 |
| universe | Extent to which the participant feels spiritually connected to the universe | 1-7 |

### Trait Emotional Intelligence

In this study, emotional intelligence (EI) was treated as a trait; a personality factor relating to various aspects of emotions. This was measured using the Short Form Trait Emotional Intelligence Questionnaire (TEIQue-SF; Petrides & Furnham, 2006). The TEIQue-SF has four subscales: wellbeing, self-control, emotionality and sociability.

| Column | Description                                                            | Values |
| ------ | ---------------------------------------------------------------------- | ------ |
| tei    | Total trait emotional intelligence score                               | 1-7    |
| wellbeing | Wellbeing | 1-7| 
| self_control   | Self-control | 1-7 |
| emotionality | Emotionality | 1-7  |
| sociability | Sociability | 1-7 |

## Predicting EI

The researchers sought to replicate previous research which found that intrinsically religious people were more spiritual than extrinsically religiously people. They also predicted a negative correlation between EI and extrinsic RO. Finally, they predicted that intrinsic RO would correlate with trait EI.

1. EI ~ age + gender + marital_status + education + religious_belief
1. EI ~ age + gender + marital_status + education + religious_belief + happiness + universe + others + nature + places
1. EI ~ age + gender + marital_status + education + religious_belief + happiness + universe + others + nature + places + ro_i + ro_e

```{r, eval=FALSE}
ei1 <- lmBF(ei ~ age + gender + marital_status + education + religious_belief, data = foo)
ei1

## RO_i

2. ro_i ~ age + gender + marital_status + education + religious_belief
2. ro_i ~ age + gender + marital_status + education + religious_belief + happiness + universe + others + nature + places
2. ro_i ~ age + gender + marital_status + education + religious_belief + happiness + universe + others + nature + places + wellbeing + self_control + emotionality + sociability + ro_e


# Ben's comments

## Use 1: Does my predictor Z explain variance, once I have controlled for other stuff?

M1 = Y ~ A + B + C….  Compared with  
M2 = Y ~ A + B + C + Z

P value from t test for Z from summary(M2)

BF = lmBF for M1 and M2 = M2/M1


## Use 2: Does this set of predictors {D,E,F} explain extra variance compared with basic predictors?
M1 = lmBF(Y ~ A + B + C)….  Compared with
M2 =   lmBF(Y ~ A + B + C + D + E + F)

F test extracted from anova(M1, M2)
BF = M2 / M1

For prediction/plotting:

M2a =   lm(Y ~ A + B + C + D + E + F)#

## Example used by Ben during this conversation

library(broom)
library(tidyverse)
library(BayesFactor)
m1f <- lm(mpg ~ wt + disp, data=mtcars)

augment(m1f) %>% ggplot(aes(wt, .fitted)) + geom_point() + geom_smooth(method='lm', se=F)
augment(m1f) %>% ggplot(aes(disp, .fitted)) + geom_point() + geom_smooth(method='lm', se=F)

m1 <- lmBF(mpg ~ wt, data=mtcars)
m2 <- lmBF(mpg ~ wt + disp , data=mtcars) # you can add factors too
m2/m1

```{r, eval=FALSE}
library(BayesFactor)
data(attitude)

## Classical regression
summary(fm1 <- lm(rating ~ ., data = attitude))

## Compute Bayes factors for all regression models
output = regressionBF(rating ~ ., data = attitude, progress=FALSE)
head(output)
## Best model is 'complaints' only

## Compute all Bayes factors against the full model, and
## look again at best models
head(output / output[63])
```

## References

Maltby, J., & Lewis, C. A. (1996). Measuring intrinsic and extrinsic orientation toward
religion: Amendments for its use among religious and non-religious samples.
Personality and Individual Differences, 21(6), 937-946.

Petrides, K. V., & Furnham, A. (2006). The Role of Trait Emotional Intelligence in a
Gender‐Specific Model of Organizational Variables 1. Journal of Applied
Social Psychology, 36(2), 552-569.

Wheeler, P., & Hyland, M. E. (2008). The development of a scale to measure the
experience of spiritual connection and the correlation between this experience
and values. Spirituality and Health International, 9(4), 193-217.
