```{r}
# summary statistics
nc_include <- nc_include %>% mutate(rank = rank(correct))
nc_summary <- nc_include %>%
  group_by(cards) %>%
  summarise(n = n(),
            median = median(correct),
            mean_rank = mean(rank),
            sum_rank = sum(rank))
nc_summary
```

**Explanation of commands:**

1. `task_by_subj %>% filter(task == 'nc')` selects rows containing data for the noun comprehension task.
1. In the preprocessed data, subjects whose data was excluded for a task have the value `NA` in the `correct` column. We exclude these rows using `drop_na()`.
1. We rank the scores in both groups using `rank(correct)` and put the rankings in `rank` using `mutate()`.
1. Finally, we group the data by `cards`, and use `summarise`, to calculate the number of children in each group `n()`, the `median()` score, `mean(rank)` and `sum(rank)`.


```{r, echo=FALSE, eval=FALSE}
# calculations for additional stats reported by SPSS
# Field(2013, p.217ff is good on this)
# z is useful for calculating r as an effect size
# See also: https://data.library.virginia.edu/the-wilcoxon-rank-sum-test/
n1 <- sum(as.integer(nc_include$cards) == 1)
wilcox.test(correct ~ cards, nc_include) %>%
    with(tibble(U = statistic,
                W = statistic + n1 * (n1 + 1) / 2,
                Z = qnorm(p.value / 2),
                p = p.value))
```
