---
title: "Preprocessing scales"
author: "Paul Sharpe, Sophie Homer"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Data required to knit
## https://github.com/ajwills72/rminr-data/tree/master/going-further/dass21.csv
## https://github.com/ajwills72/rminr-data/tree/master/going-further/sses.csv
##
## I check out rminr-data and make a symbolic link to going-further

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
options(tibble.width = Inf) # show all columns in output
library(kableExtra)
library(pander)
```

# Contents

- [Introduction](#intro)

- [Getting started](#start)

- [Selecting data](#select)

- [Handling missing data](#missing)

- [Calculating subscale scores](#subscales)

- [Tidying survey data](#tidy)

- [Reverse scoring survey items](#reverse)

<a name="intro"></a>

# Introduction

Intelligence, personality, and many other psychological constructs are often measured using scales. This type of data is normally collected using questionnaires (also called surveys). Answers to the questions are given numerical values, most commonly using a Likert scale. Likert scales associate numbers with a set of answers which express some degree of agreement each question e.g. `0=Not at all, 1=A little, 2=Somewhat, 3=A lot, 4=Extremely`. A formula is applied to the scores for questions which make up the construct to calculate an overall number for the scale. This could be all questions, or a subset for scales which contain subscales. The formula often just consists of adding up the individual scores (more on this below).

A psychometric scale is a scale which has undergone some degree of testing to ensure that it is valid and reliable. A valid scale is a true measure of the underlying construct, e.g. it measures intelligence, rather than something else. A reliable scale gives consistent results, i.e. a person who completed the scale at different times would produce similar scores, as would two people who are similar in terms of the construct measured by the scale. Most published scales have been tested to ensure they are valid and reliable, so it's advisable to use an existing scale if one exists, before creating your own.

Surveys can be created using [JISC](https://www.jisc.ac.uk/), [Gorilla Survey](https://gorilla.sc/), [OpenSesame](https://osdoc.cogsci.nl/), [The Experiment Factory](https://expfactory.github.io/experiments/), [Qualtrics](https://www.qualtrics.com/) and many other software packages. Most software will allow you to save your data as a CSV file. The precise structure of the data varies between packages, so you are likely to have to start by [preprocessing](preproc.html) your data.

In this worksheet, we'll cover some common techniques which you are likely to use to preprocess psychometric scale data. These techniques should be useful regardless of the software you used to collect your survey data, although they will need slight modifications depending on the way your raw data is organised.

<a name="start"></a>

# Getting started

To prepare for this worksheet, follow these steps:

1. Create a new R project.

1. In your project, create a folder named `going-further` 

1. You will find the files `dass21.csv` and `state-self-esteem.sav` in the `going-further` folder in git repository we used [previously](preproc.html#load). Upload these files into the `going-further` folder you just created.

1. Create a script named `scales.R` and add the code in this worksheet to this script as you work through each section.

We start with some lines to clear the workspace and load `tidyverse`.

```{r init-load, message=FALSE}
rm(list = ls()) # clear the environment
library(tidyverse)
```

<a name="select"></a>

# Selecting data

Our first step will be to remove columns from the raw survey data which aren't needed for data analysis.  We'll demonstrate this using some real data from the Depression Anxiety Stress Scales—21 (DASS-21, Heatherton & Polivy, 1991). The DASS-21 is a 21-item scale for measuring depression, anxiety and stress. This data is stored in the file `dass21.csv` which you added to your project above.

```{r pre-process-dass}
dass21_raw <- read_csv("going-further/dass21.csv")
dass21_raw <- select(dass21_raw, partID, Age:DASS21)
```

**Explanation of command:**

1. We read the CSV file using `read_csv`, creating the data frame `dass21_raw`.

1. We then `select()` just the columns in `dass21_raw` that we want to keep. The first column we `select()`, is the participant ID, which is stored in the `partID` column. Arguments to `select()` can also be consecutive ranges of columns in a data frame, consisting of the first and last column name (ordered from left to right), separated by a `:`. This avoids having to type out long lists of column names.  Here we use `Age:DASS21` to select all columns between `Age` and `DASS21`.

The table below shows the first few rows from `dass21_raw`. In this study, the data was recorded in "wide" format, i.e. one row for each participant. Notice that our data frame contains only the columns that we selected in the commands above. The DASS-21 scores are in columns `DASS1`-`DASS21`.

```{r dass21, echo=FALSE}
head(dass21_raw) %>% pander(split.table = Inf)
```

<a name="missing"></a>

# Handling missing data

If participants don't complete (or partially complete) a survey, you may want to exclude their data from your analyses. If we look at a section of the DASS-21 data, we can see that data is missing for some participants.

```{r missing, echo=FALSE}
slice(dass21_raw, 73:76) %>% pander(split.table = Inf)
```

Participants `108`, and `109` (rows 3 and 4) have numbers in all columns, indicating that their data is complete. However, participants `106` and `107` have cells containing the value `NA`, which means these cells in the CSV file were empty. For participant `106`, all cells are `NA` (perhaps they dropped out of the study), and for participant `107`, all of the DASS-21 cells are `NA` (perhaps they skipped this survey).

We can exclude any rows with missing data as follows:

```{r exclude-missing}
dass21 <- dass21_raw %>% drop_na()
```

**Explanation of command:**

1. We pipe the raw data into the `drop_na()` function (part of the `tidyverse` package). This drops any rows with a column containing `NA`, which includes data for participants `106` and `107`.

```{r exclude, echo=FALSE}
slice(dass21, 68:72) %>% pander(split.table = Inf)
```

<a name="subscales"></a>

# Calculating subscale scores

Our next step is to calculate the scores for the constructs measured by our scale. Many scales consist of groups of questions which measure multiple, distinct constructs. The DASS-21 is an example of a scale with subscale scores for depression, anxiety and stress. These are calculated by adding together responses for specific items, which we can do using the `rowSums()` function:

```{r dass21-subscales}
dass21 <- mutate(dass21, depression = rowSums(dass21[c(4+3,4+5,4+10,4+13,4+16,4+17,4+21)]))
dass21 <- mutate(dass21, anxiety = rowSums(dass21[c(4+2,4+4,4+7,4+9,4+15,4+19,4+20)]))
dass21 <- mutate(dass21, stress = rowSums(dass21[c(4+1,4+6,4+8,4+11,4+12,4+14,4+18)]))
```

**Explanation of commands:**

1. We use `mutate()` to create a `depression` column which is the sum of items 3, 5, 10, 13, 16, 17 and 21. Item 1 of the DASS-21 data is in column 5 of `dass21`, so we add 4 to each item number to select the correct columns to add together. We assign the result back to `dass21` to create a `depression` column for each row.

1. We use similar commands to create `anxiety` and `stress` columns based on the items which make up these subscales.

Our `dass21` data frame now looks has a depression, anxiety and stress score for each participant:

```{r, echo=FALSE}
dass21 %>% head(3) %>% pander(split.table = Inf)
```

<a name="tidy"></a>

# Tidying survey data

Some data benefits from a little more tidying than simply removing columns which aren't required. We'll demonstrate this more advanced preprocessing the other dataset (`sses.csv`) which you added to your project above.This data comes from from an experiment in which self-esteem was measured before and after participants completed one of two mental imagery conditions, or a control condition. This experiment used the State Self-Esteem Scale (SSES, Osman et al., 2012), a 20-item scale used to measure short-lived (state) changes in self-esteem.

```{r, R.options = list(width = 100)}
sses <- read_csv('going-further/sses.csv')
```

```{r, echo=FALSE}
sses %>% head(3) %>% pander(split.table = Inf)
```

The data will be easier to analyse if rename the columns. It will also be useful to divide the data into two data frames, one for the pre-intervention SSES, the other for the post-intervention SSES. We can do all of this in a single pipeline:

```{r preprocess-sses-pre}
sses_pre_raw  <- sses %>% select(1, 5:25) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition), time = factor('pre')) %>%
  select(subj, condition, time, q1:q20)
```

**Explanation of pipeline commands:**

1. `sses_pre_raw  <- sses %>% select(1, 5:25)` - We save the output of the pipeline in `sses_pre_raw`. First, we pipe the `sses` data frame into a `select` which keeps column 1, and columns 5:25. Column 1 is the participant id, columns 5:24 are the SSES scores, and column 25 is the participant's condition.

1. `set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q"))` - Next, we use the function `set_names()` to rename our columns. The `~` is a way of telling `set_names()` to apply a function each column name. The remainder of the command is a "sub-pipeline" which tidies up the column name. The command `str_to_lower(.)` converts a string (the `.` means the current column name) to lower case. This lower case name is piped to `str_replace_all("pre_sse_", "q"))` which replaces any string with the prefix `pre_sse_` with `q`. All our columns are now lowercase, and the SSES questions are named `q1:q20`.

1. `mutate(subj = factor(partid), condition = factor(condition), time = factor('pre'))` - Next, we use `mutate` to add and modify some columns. The argument `subj = factor(partid)` creates a new column named `subj` (which is a bit clearer than `partid`) by copying the `partid` column and making it a factor. The argument `condition = factor(condition)` makes the `condition` column a factor. The argument `time = factor('pre')` creates a new factor called `time` and sets all values to `pre`.

1. `select(subj, condition, time, q1:q20)` - Finally, we use `select()` to put the columns into a convenient order for viewing.

We now have our pre-intervention SSES data in a data frame with clearer column headings:

```{r, echo=FALSE}
sses_pre_raw %>% head(3) %>% pander(split.table = Inf)
```

We use a similar pipeline to create a tidy data frame containing the post-intervention SSES data.

```{r preprocess-sses-post}
sses_post_raw <- sses %>% select(1, 25:45) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("post_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition), time = factor('post')) %>%
  select(subj, condition, time, q1:q20)
```

**Explanation of command:**

1. This is indentical to the previous pipeline, except that the post-intervention SSES data are in columns `26:45`, these columns have the prefix `post_sse_` rather than `pre_sse_`, and we use the value `post` in our `time` factor.

```{r, echo=FALSE}
sses_post_raw %>% head(3) %>% pander(split.table = Inf)
```

<a name="reverse"></a>

# Reverse scoring survey items

The SSES questions are scored from `0 (Not at all)` to `4 (Extremely)`. Question 1 reads `I feel confident about my abilities`, so a higher score indicates higher self-esteem. However, a number of questions on the SSES are 'reverse scored'. For example, question 2 reads `I am worried about whether I am regarded as a success or failure`, so in this case a higher score indcates lower self-esteem. Surveys often use reverse scored questions to encourage participants to pay attention to the wording of each question. Before calculating a total SSES score, we need to reverse the reverse scored items, so that a higher number means higher self-esteem for all items.  

## Writing a function to reverse score survey items

We have a number of reverse scored items which we need to reverse for each participant. It would be laborious to write code to do this for each one, so we'll write some reusable code which can reverse any reverse coded score. This requires a brief digression into an explanation of functions.

Functions are reusable lines of code which perform a particular function (hence the name), and are one of the most powerful features of programming languages such as R. At this point you will be familiar with _using_ functions. For example `rm()` is a function for removing objects from the R environment, and is part of the base `R` language. The `rm()` function is reusable in the sense that you can use the same function to do slightly different things. For example, `rm('foo')` would remove the single object `foo` from your environment. The string `foo` is called an _argument_ and is the data processed by `rm()`. The same function can also process a list of variable names, so calling `rm(ls())`, first calls `ls()` which lists all objects in your environment, and passes the results as an argument to `rm()` which removes them, thereby cleaning your environment of all objects. Similarly, you've used `read_csv()`, with different filename arguments to read different data files. The `read_csv()` function is part of the `readr` package which is loaded when you call `library(tidyverse)` (`library()` is another base `R` function). 

Removing objects from the environment and reading files are such common tasks that they are, respectively, part of base `R` and a package that you can load using `library()`. However, we can use a special base `R` function called `function()`, to define our own functions for more specific tasks that we want to do repeatedly. This is not particularly complicated, but requires a little explanation if you've never encountered user-defined functions before (you can find more details in [the R manual](https://cran.r-project.org/doc/manuals/r-release/R-intro.html#Writing-your-own-functions)). We'll explain the main steps in defining a function, by writing a simple function to reverse survey items which are reverse coded.

```{r reverse-score-survey}
# Reverse score survey items
# @param scores raw scores data frame
# @param reverse character vector of columns to be reverse scored
# @param max integer value of maximum score
reverse_score_survey <- function(scores, reverse, max) {
  reverse_score <- function(score) { max - score } # for scores starting at 1, max + 1 - score
  scores %>% mutate_at(reverse, reverse_score)
}
```

**Explanation of commands:**

1. Just like functions we use, our function consists of some reusable code, and the data which it processes. We create our function using the `function()` function. The arguments that our function accepts, `scores`, `reverse`, and `max` go inside the parentheses `()`. The code goes inside the curly brackets `{` and `}`. We name our function by assigning the result of `function(...) { ... }` to the variable `reverse_score_survey`.

1. Our `reverse_score_survey()` function requires three arguments: `scores` is a data frame containing raw survey responses, i.e. with some items reverse scored, `reverse` is a character vector containing the column names in `scores` that are reverse scored, and `max` is the maximum score available for an item on the Likert scale used for each item.

1. Inside `reverse_score_survey()`, we define another function called `reverse_score()`. Defining one function within another function is perfectly allowable in R. The `reverse_score()` function takes a `score` argument, and reverses it using the calculation `max - score`. For example, `reverse_score(4)`, would return the value `0` (`4 - 4`).  If your survey scores began at 1, you could adjust the calculation to `max + 1 - score`.

1. The final line in `reverse_score_survey()` pipes our raw scores into `mutate_at(reverse, reverse_score)`. The `mutate_at()` function applies a function to a subset of columns in a data frame. The first argument `reverse` is the list of column names to be reversed. The second argument `reverse_score` is the name of the function we wrote to reverse the score.

1. A function returns the last object it creates to the code that called it. In `reverse_score_survey()`, this will be a data frame with all scores adjusted to be from 0 to `max`.

This was quite a long explanation for a simple, but very powerful four lines of code. By calling this function with different arguments, we can reverse the reverse coded items on any survey in which questions are answered using a Likert scale. To demonstrate this, we can now call `reverse_score_survey()` to reverse score our pre and post intervention SSES scores:

```{r reverse-pre}
reverse_questions <- c('q2', 'q4', 'q5', 'q7', 'q8', 'q10', 'q13', 'q15', 'q16', 'q17', 'q18',
                       'q19', 'q20')
sses_pre  <- reverse_score_survey(sses_pre_raw, reverse_questions, 4)
sses_post <- reverse_score_survey(sses_post_raw, reverse_questions, 4)
```

**Explanation of commands:**

1. We set `reverse_questions` to be a vector of the column names of the SSES items that are reverse scored.

1. Then we call `reverse_score_survey(sses_pre_raw, reverse_questions, 4)`. When the function is run, `sses_pre_raw` is assigned to the `scores` argument, `reverse_questions` to the `reverse` argument, and `4` to the `max` argument. The data frame returned by `reverse_score_survey()` (higher score meaning higher self-esteem for all items) is assigned to `sses_pre`.

1. Similarly, we call the same function with `sses_post_raw` to reverse score the post-intervention surveys and assign the result to `sses_post`.

Now that our items are all scored in the same direction, we can calculate the total SSES baseline scores using the same technique we used for [calculating subscale scores](#subscales):

```{r pre-score, R.options = list(width = 100)}
sses_pre <- mutate(sses_pre, total = rowSums(sses_pre[4:23]))
sses_post <- mutate(sses_post, total = rowSums(sses_post[4:23]))
```

```{r, echo=FALSE}
sses_pre %>% head(3) %>% pander(split.table = Inf)
```

# References

Heatherton, T. F., & Polivy, J. (1991). Development and validation of a scale for measuring state self-esteem. _Journal of Personality and Social Psychology, 60(6)_, 895.

Osman, A., Wong, J. L., Bagge, C. L., Freedenthal, S., Gutierrez, P. M., & Lozano, G. (2012). The Depression Anxiety Stress Scales—21 (DASS-21): Further Examination of Dimensions, Scale Reliability, and Correlates. _Journal of Clinical Psychology, 68(12)_, 1322–1338. https://doi.org/10.1002/jclp.21908

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

