---
title: "Data preprocessing for scales"
author: "Paul Sharpe, Andy Wills, Sophie Homer"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Data required to knit
## https://github.com/ajwills72/rminr-data/tree/master/going-further/dass21.csv
## https://github.com/ajwills72/rminr-data/tree/master/going-further/sses.csv
##
## I check out rminr-data and make a symbolic link to going-further

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and output.
knitr::opts_chunk$set(echo = TRUE, comment = NA, cache = TRUE)
options(tibble.width = Inf) # show all columns in output
library(kableExtra)
library(pander)
```

## Contents

- [Introduction](#intro)

- [Getting started](#start)

- [Loading and selecting data](#select)

- [Handling missing data](#missing)

- [Calculating subscale scores](#subscales)

- [Exercise 1](#ex1)

- [Tidying survey data](#tidy)

- [Exercise 2](#ex2)

<a name="intro"></a>

## Introduction

Intelligence, personality, and many other psychological constructs are often measured using scales. This type of data is normally collected using questionnaires (also called surveys). Answers to the questions are given numerical values, most commonly using a Likert scale. Likert scales associate numbers with a set of answers which express some degree of agreement with each question e.g. `0=Not at all, 1=A little, 2=Somewhat, 3=A lot, 4=Extremely`. A formula is applied to the scores for some or all of the questions to calculate an overall score for the scale. The formula often just consists of adding up the individual scores (more on this below). This worksheet assumes that your survey software has recorded Likert responses as numbers. Refer back to the [Cleaning up questionnaire data worksheet](https://benwhalley.github.io/rmip/data.html#using-recode) if you need a reminder of how to convert text responses to numbers.

A psychometric scale is a scale which has undergone some degree of testing to ensure that it is a valid and reliable measure of the underlying construct. For example, a valid intelligence scale would truly measure intelligence, rather than some other construct (e.g. memory). A reliable scale gives consistent results, i.e. a person who completed the scale at different times would produce similar scores, as would two people who are similar in terms of the construct measured by the scale. Most published scales have been tested to ensure they are valid and reliable, so it's advisable to use an existing scale if one exists, before creating your own.

Surveys can be created using [JISC](https://www.jisc.ac.uk/), [Gorilla Survey](https://gorilla.sc/), [OpenSesame](https://osdoc.cogsci.nl/), [The Experiment Factory](https://expfactory.github.io/experiments/), [Qualtrics](https://www.qualtrics.com/) and many other software packages. Most software will allow you to save your data as a CSV file. The precise structure of the data varies between packages, so you are likely to have to start by [preprocessing](preproc.html) your data.

In this worksheet, we'll cover some common techniques you are likely to use to preprocess psychometric scale data. These techniques should be useful regardless of the software you used to administer your survey data, although they will need slight modifications depending on the way your raw data is organised.

<a name="start"></a>

## Getting started

To prepare for this worksheet:

1. Open the `rminr-data` project we used [previously](preproc.html#load).

1. If you don't see a folder named `going-further`, it means you created your project _before_ the data required for this worksheet was added to the `rminr-data` git repository. You can get the latest files by asking git to "`pull`" the repository. Select the `Git` tab, which is located in the row of tabs which includes the `Environment` tab. Click the `Pull` button with a downward pointing arrow. A window will open showing the files which have been pulled from the repository. Close the `Git pull` window.

1. Open the `Files` tab. The `going-further` folder should contain the files `dass21.csv` and `sses.sav`.

1. Create a script named `scales.R` in the `rminr-data` folder (the folder above `going-further`). Add the comments and code to this script as you work through each section of the worksheet.

We start with some lines to clear the workspace and load `tidyverse`.

**Enter these comments and commands into your script, and run them:**

```{r init-load, message=FALSE}
# Data preprocessing for scale
# Clear the environment
rm(list = ls()) 
# Load tidyverse
library(tidyverse)
```

<a name="select"></a>

## Loading and selecting data

Our first step will be to load the data and remove columns from the raw survey data which aren't needed for analysis. We'll demonstrate this using some real data from the Depression Anxiety Stress Scales—21 (DASS-21, Henry & Crawford, 2005), a 21-item scale for measuring depression, anxiety and stress.

**Enter these comments and commands into your script, and run them:**

```{r pre-process-dass, message=FALSE}
# Load data
dass21_raw <- read_csv("going-further/dass21.csv")
# Select relevant columns of data
dass21_raw <- dass21_raw %>% select(partID, Age:DASS21)
```

**Explanation of command:**

1. We read the DASS-21 CSV file into the data frame `dass21_raw`.

1. We then `select()` just the columns in `dass21_raw` that we want to keep. The first column we `select()`, is the participant ID, which is stored in the `partID` column. Arguments to `select()` can also be consecutive ranges of columns in a data frame, consisting of the first and last column name (ordered from left to right), separated by a `:`. This avoids having to type out long lists of column names.  Here we use `Age:DASS21` to select all columns between `Age` and `DASS21`.

The table below shows the first few rows from `dass21_raw`. In this study, the data was recorded in "wide" format (one row for each participant). Notice that our data frame contains only the columns that we selected in the commands above. The DASS-21 scores are in columns `DASS1`-`DASS21`.

```{r dass21, echo=FALSE}
head(dass21_raw) %>% pander(split.table = Inf)
```

<a name="missing"></a>

## Handling missing data

If participants don't complete (or partially complete) a survey, you may want to exclude their data from your analyses. Here are some rows from `dass21_raw`.

```{r missing, echo=FALSE}
slice(dass21_raw, 73:76) %>% pander(split.table = Inf)
```

We can see that participants `108`, and `109` (rows 3 and 4) have numbers in all columns, indicating that their data is complete. However, participants `106` and `107` have cells containing the value `NA`, which means these cells in the CSV file were empty. For participant `106`, all cells are `NA` (perhaps they dropped out of the study), and for participant `107`, all of the DASS-21 cells are `NA` (perhaps they skipped this survey).

If you select `dass21_raw` in the **Environment** pane and look through the rest of the data, you'll see that participants 35, 49, 61, and 77 also have no data for this survey. We exclude these participants from the data frame.

**Enter these comments and commands into your script, and run them:**

```{r exclude-missing}
# Exclude participants with no data
exclude <- c(35,49,61,77,106,107)
dass21 <- dass21_raw %>% filter(!(partID %in% exclude))
```

**Explanation of commands:**

Line 1 creates a list of the participant numbers we wish to exclude. In line 2, we remove those participants from `dass21_raw`. The `filter` command will be familiar from previous worksheets. The command `partID %in% exclude` means 'any participant whose subject number is in our list `exclude`'. The use of `!()` in the filter statement means `not`. So, `filter(!(partID %in% exclude))` means keep the participants whose subject number is _not_ in the `exclude` list. 

If you look at the **Value** column in the **Environment** pane, you'll see that `dass21` now has six fewer rows than `dass21_raw`.

<a name="subscales"></a>

## Calculating subscale scores

Our next step is to calculate the scores for the constructs measured by our scale. Many scales consist of groups of questions which measure multiple, distinct constructs. The DASS-21 is an example of a scale with subscale scores for depression, anxiety and stress. These are calculated by adding together responses for specific items, which we can do using the `rowSums()` function.

**Enter these comments and commands into your script, and run them:**

```{r dass21-subscales}
# Calculate depression subscale score
dass21 <- dass21 %>%
  mutate(depression = rowSums(dass21[4 + c(3,5,10,13,16,17,21)]))
# Add relevant columns to 'dass21_total'
dass21_total <- dass21 %>% select(partID, Age, Gender, depression)
```

**Explanation of command:**

* We use `mutate()` to create a `depression` column which is the sum of items 3, 5, 10, 13, 16, 17 and 21. Item 1 of the DASS-21 data is in column 5 of `dass21`, so we add 4 to each item number to select the correct columns to add together. The command `dass21[4 + c(3,5,10,13,16,17,21)]` is an example of "vectorised addition". It adds `4` to each of the columns defined in the vector to the right of the `+`. For each row, the values in the resulting columns are added together using `rowSums()`. We assign the result back to `dass21`, thereby creating a `depression` column for each row.

<a name="ex1"></a>

## Exercise 1

Use similar commands to add scores for anxiety and stress to `dass21`. The anxiety subscale is the sum of questions 2,4,7,9,15,19 and 20. The stress subscale is the sum of questions 1,6,8,11,12,14 and 18. After running your commands, the first few rows of `dass21_total` should look like this:

```{r ex1, echo=FALSE}
dass21 <- mutate(dass21, anxiety = rowSums(dass21[4 + c(2,4,7,9,15,19,20)]))
dass21 <- mutate(dass21, stress = rowSums(dass21[4 + c(1,6,8,11,12,14,18)]))
dass21_total <- select(dass21, partID, Age, Gender, depression, anxiety, stress)
```

```{r ex1-correct, echo=FALSE}
dass21_total %>% head(10) %>% pander()
```

**Copy the R code you used for this exercise, along with appropriate comments, into PsycEL.**

<a name="tidy"></a>

## Tidying survey data

Some data benefits from a little more tidying than simply removing columns which aren't required. We'll demonstrate this more advanced preprocessing using a different dataset. This data came from from an experiment in which self-esteem was measured before and after participants completed one of two mental imagery conditions, or a control condition.

 * Condition 1: participants visualised a negative mental image of themself
 * Condition 2: participants visualised a negative mental image of someone else
 * Condition 0 (control): participants did a card sorting task and did not think of any images.

The experiment used the State Self-Esteem Scale (SSES, Heatherton & Polivy, 1991), a 20-item scale used to measure short-lived (state) changes in self-esteem.

**Enter this comment and command into your script, and run it:**

```{r, R.options = list(width = 100), message=FALSE}
# Load data into 'sses'
sses <- read_csv('going-further/sses.csv')
```

```{r, echo=FALSE}
sses %>% head(3) %>% pander(split.table = Inf)
```

The data will be easier to analyse if we rename the columns. It will also be useful to divide the data into two data frames, one for the pre-intervention SSES, the other for the post-intervention SSES. We'll do this in stages. 

**Enter this comment and command into your script, and run it:**

```{r preprocess-sses-pre-1}
# Place pre-intervention SSES into 'sses_pre_raw'
sses_pre_raw  <- sses %>% select(1, 5:25)
```

**Explanation of command:**

* `sses_pre_raw  <- sses %>% select(1, 5:25)` - We `select()` column 1, and columns 5:25 from `sses`, and store the resulting data frame in `sses_pre_raw`. Column 1 is the participant id, columns 5:24 are the SSES scores, and column 25 contains a number indicating which of the three experimental conditions the subject was assigned to.

Here are the first three participants of our pre-intervention data:

```{r echo=FALSE}
sses_pre_raw %>% head(3) %>% pander(split.table = Inf)
```

Next, we'll rename the SSES columns based on their question number. This will make them easier to refer to in the rest of our code.

**Enter this comment and command into your script, and run them:**

```{r preprocess-sses-pre-2}
# Rename columns
sses_pre_raw  <- sses_pre_raw %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q"))
```

**Explanation of command:**

1. `set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q"))` - We use the function `set_names()` to rename our columns. The `~` is a way of telling `set_names()` to apply a function to each column name. The remainder of the command is a "sub-pipeline" which tidies up the column name. The command `str_to_lower(.)` converts a string (the `.` means the current column name) to lower case. This lower case name is piped to `str_replace_all("pre_sse_", "q"))` which replaces any string with the prefix `pre_sse_` with `q`. All our columns are now lowercase, and the SSES questions are named `q1:q20`.

```{r echo=FALSE}
sses_pre_raw %>% head(3) %>% pander(split.table = Inf)
```

Now we'll convert some columns to [factors](anova1.html#anovaWS).

**Enter this comment and command into your script, and run them:**

```{r preprocess-sses-pre-3}
# Convert columns to factors; add factor column 'time', set to 'pre'; select relevant columns
sses_pre_raw <- sses_pre_raw %>%
  mutate(subj = factor(partid), condition = factor(condition),
         time = factor('pre')) %>%
  select(subj, condition, time, q1:q20)
```

1. `mutate(subj = factor(partid), condition = factor(condition), time = factor('pre'))` -  We use `mutate` to add and modify some columns. The argument `subj = factor(partid)` creates a new column named `subj` (which is a bit clearer than `partid`) by copying the `partid` column and making it a factor. The argument `condition = factor(condition)` makes the `condition` column a factor. The argument `time = factor('pre')` creates a new factor called `time` and sets all values to `pre`.
1. `select(subj, condition, time, q1:q20)` just puts our columns in a more logical order.

Our data is now much tidier:

```{r echo=FALSE}
sses_pre_raw %>% head(3) %>% pander(split.table = Inf)
```

Note that we could do all of these steps in a single pipeline (**DO NOT enter these commands**, you do not need to do the same thing twice, this is just an illustration of how the previous commands could be combined).
 
```{r preprocess-sses-pre}
sses_pre_raw  <- select(sses, 1, 5:25) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("pre_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition),
         time = factor('pre')) %>%
  select(subj, condition, time, q1:q20)
```

<a name="ex2"></a>

## Exercise 2

Write a similar pipeline (including comments) to create a data frame named `sses_post_raw` containing the post-intervention SSES data. The condition and post-intervention SSES data are in columns `25:45`. The SSES columns have the prefix `post_sse_` rather than `pre_sse_`. Set the value in the `time` factor to `post`. After running your commands, the first few rows of `sses_post_raw` should look like this:

```{r ex2, echo=FALSE}
sses_post_raw <- select(sses, 1, 25:45) %>%
  set_names(~ str_to_lower(.) %>% str_replace_all("post_sse_", "q")) %>%
  mutate(subj = factor(partid), condition = factor(condition),
         time = factor('post')) %>%
  select(subj, condition, time, q1:q20)
```

```{r ex2-correct, echo=FALSE}
sses_post_raw %>% head(3) %>% pander(split.table = Inf)
```

**Copy the R code and comments you used for this exercise into PsycEL.**

## References

Heatherton, T. F., & Polivy, J. (1991). [Development and validation of a scale for measuring state self-esteem](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.476.4746&rep=rep1&type=pdf) _Journal of Personality and Social Psychology, 60(6)_, 895.

Henry, J. D., & Crawford, J. R. (2005). [The short-form version of the Depression Anxiety Stress Scales (DASS-21): Construct validity and normative data in a large non-clinical sample](https://onlinelibrary.wiley.com/doi/pdf/10.1348/014466505X29657?casa_token=qL449BGkw4EAAAAA:qrRVqngwUjujoSuG3WwppU0Cma7Kv0YD03eAy6xrmIkLFpyr3mOv7bMv-LiWkpS46Cqr7rgDay5b-A) British Journal of Clinical Psychology, 44(2), 227–239.

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

