---
title: "Analysing survey data"
author: "Paul Sharpe, Sophie Homer"
output: html_document
---

```{r setup, include=FALSE}
## DEVELOPERS: Uncomment one option, as appropriate

## Show only commands.
## knitr::opts_chunk$set(echo = TRUE, message = FALSE, results='hide', fig.keep = 'none', comment=NA)

## Show commands and putout.
knitr::opts_chunk$set(echo = TRUE, comment=NA, cache = TRUE)
```

# Contents

- [Introduction](#intro)

- [Getting started](#start)

- [TODO: Reverse score some survey items](#reverse)

- [TODO: Calculate a total score for pre and post](#score)

- [TODO: Calculate cronbach’s alpha](#cronbach)

- [TODO: Compare baseline SSES scores between conditions using a one-way ANOVA (hoping not to find differences at baseline)](#baseline)

- [# TODO: Run a within-between ANOVA with time as the repeated measures factor and condition as the between-subjects factor](#prepost)

<a name="intro"></a>

# Introduction

In this worksheet, we'll analyse data from a survey called the State Self-Esteem Scale (SSES). The SSES consists of 20 questions, each of which is scored from 0 (Not at all) to 4 (Extremely). 

<a name="start"></a>

# Getting started

We'll be using data from an experiment which explored mental imagery and self-esteem. Create a new R project and upload the file `state-self-esteem.sav`, which you will find in the git repository we used [previously](preproc.html#load).

FIXME <span style="color: red;">Add state-self-esteem.sav to rminr-data</span>

```{r init-load, message=FALSE} 
library(tidyverse)
# TODO: Should we leave this as SPSS to teach read.spss()?
library(foreign)
sses <- read.spss(paste0('state-self-esteem.sav'), to.data.frame=TRUE)
```

## Pre-processing

```{r pre-process}
sses_pre_raw  <- sses %>% select(1, 5:25) %>%
    set_names(~ str_to_lower(.) %>%
                  str_replace_all("pre_sse_", "q")) %>%
  mutate(condition = factor(condition))
sses_pre_raw$condition <- recode_factor(sses_pre_raw$condition, `0` = 'control',
                                        `1` = 'self', `2` = 'other')

sses_post_raw <- sses %>% select(1, 25:45) %>%
      set_names(~ str_to_lower(.) %>%
                  str_replace_all("post_sse_", "q")) %>%
  mutate(condition = factor(condition))
sses_post_raw$condition <- recode_factor(sses_post_raw$condition, `0` = 'control',
                                        `1` = 'self', `2` = 'other')

```

<a name="reverse"></a>

# Reverse score survey

Explain reverse coding.

Some surveys have some reverse code items because...


```{r reverse}
# Reverse score survey items
# @param scores raw scores
# @param reverse character vector of columns to be reverse scored
# @param max integer value of maximum score
reverse_score_survey <- function(scores, reverse, max) {
  reverse_score <- function(score) { max - score } # for scores which start at 0, for starting at 1, max + 1 - score
  scores %>% mutate_at(reverse, reverse_score)
}

reverse_questions <- c('q2', 'q4', 'q5', 'q7', 'q8', 'q10', 'q13', 'q15', 'q16', 'q17', 'q18',
                       'q19', 'q20')
sses_pre <- sses_pre_raw %>%
  group_by(condition) %>%
  group_modify(~ reverse_score_survey(.x, reverse_questions, 4)) %>%
  ungroup()
sses_post <- sses_post_raw %>%
  group_by(condition) %>%
  group_modify(~ reverse_score_survey(.x, reverse_questions, 4)) %>%
  ungroup()
```

<a name="score"></a>

# Calculate a total score for pre and post

```{r score}
sses_pre  <- sses_pre  %>% mutate(total = rowSums(.[3:22]))
sses_pre %>%
  group_by(condition) %>%
  summarise(mean = mean(total), sd = sd(total))

sses_post <- sses_post %>% mutate(total = rowSums(.[3:22]))
sses_post %>%
  group_by(condition) %>%
  summarise(mean = mean(total), sd = sd(total))
```

<a name="cronbach"></a>

# Calculate Cronbach’s alpha

```{r cronbach}
library(psy)
sses_cronbach <- function(scores) {
  cronbach <- cronbach(select(scores,q1:q20))
  data.frame(cronbach$alpha)
}

sses_pre_cronbach <- sses_pre %>%
  group_by(condition) %>%
  group_modify(~ sses_cronbach(.x))
sses_pre_cronbach

sses_post_cronbach <- sses_post %>%
  group_by(condition) %>%
  group_modify(~ sses_cronbach(.x))
sses_post_cronbach
```

<a name="baseline"></a>

# Compare baseline SSES scores between conditions

* one-way ANOVA
* hope not to find differences

<a name="prepost"></a>

# Pre-post

* Within-between ANOVA sses ~ time*condition
* time is the repeated measures factor
* condition is the between-subjects factor

___

This material is distributed under a [Creative Commons](https://creativecommons.org/) licence. CC-BY-SA 4.0. 

